---
title: "Filter scRNA-Seq and VDJ data"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose: 
* annotate cell types and define expanded clonotypes (B cells and T cells)
  * QC filtering, Doublet removal and automated cell type annotation was performed in **0_preprocessing.Rmd**
* overlap single cell RNA-Seq and VDJ data on proceed with analyzing overlapping cells

## Prior to running this code:
* Before starting set working directory to the **CompImmunologyWorkshopUCSF** folder
  * setwd(**"[PATH]/CompImmunologyWorkshopUCSF"**)

```{r,warning=FALSE,message=FALSE}
# Set Working Directory before starting

# Immcantation
suppressPackageStartupMessages(library(alakazam))
suppressPackageStartupMessages(library(shazam))
suppressPackageStartupMessages(library(tigger))
suppressPackageStartupMessages(library(dowser))
suppressPackageStartupMessages(library(airr))

suppressPackageStartupMessages(library(glmGamPoi))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(dittoSeq))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(devtools))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(impactSingleCellToolkit))

# Immcantation Results
bcr_heavy_fh <- "data/results_immcantation/BCR_CSFPB_heavy_germ-pass.tsv"
bcr_light_fh <- "data/results_immcantation/BCR_CSFPB_light_germ-pass.tsv"
tcr_beta_fh  <- "data/results_immcantation/TCR_CSFPB_heavy_germ-pass.tsv"
tcr_alpha_fh <- "data/results_immcantation/TCR_CSFPB_light_germ-pass.tsv"

# seed number
seed.number <- 12345

# Workshop specific functions
source("resources/functions_workshop.R")
```


# 1. Load RNA-Seq object

* At this stage of the analysis scRNA-Seq, BCR and TCR repertoire data have been processed and overlapped
* Doublet were removed computationally using 'DoubletFinder'
* cells with more than 10% mitochondrial gene expression were omitted
* Only B cells and T cells exist in this object

```{r}
rnaseq_obj_full <- readRDS("objects/seurat.obj.processed.rds")
```

# 2. Load and format TCR/BCR repertoire results

* Clonotype definitions for BCR are based on the Heavy chain only
* Clonotype definitions for TCR include both Alpha and Beta chains

### a. Load immcantations results
```{r}
# TCR
db.tra   <- read_rearrangement(tcr_alpha_fh) %>% as.data.frame()
db.trb   <- read_rearrangement(tcr_beta_fh) %>% as.data.frame()
# BCR
db.heavy <- read_rearrangement(bcr_heavy_fh) %>% as.data.frame()
db.light <- read_rearrangement(bcr_light_fh) %>% as.data.frame()
```

### b. filter contigs and merge all chains
* All contigs are in frame
```{r}
bcr_df <- mergeVDJresults(db.heavy,db.light,umi.thresh = 3,assay = "bcr")
tcr_df <- mergeVDJresults(db.trb,db.tra,umi.thresh = 2, assay = "tcr")

# format VDJ unique cell IDs to match scRNA-Seq (function only for this workshop)
formatUniqueIDcolum <- function(uni_id_column,sep = ":") {
  samples <- tstrsplit(uni_id_column, sep)[[1]]
  cells   <- tstrsplit(uni_id_column, sep)[[2]]
  fmt_ids <- paste(cells,samples,sep = "-")
  
  return(fmt_ids)
}

bcr_df$unique_cell_id <- formatUniqueIDcolum(bcr_df$unique_cell_id)
tcr_df$unique_cell_id <- formatUniqueIDcolum(tcr_df$unique_cell_id)
```


# 3. Only analyze cells shared across all datasets

### a. get vector of overlapping cells
```{r}
cells_rnaseq      <- colnames(rnaseq_obj_full)
cells_vdj         <- unique(c(bcr_df$unique_cell_id,
                              tcr_df$unique_cell_id))
overlapping_cells <- cells_rnaseq[cells_rnaseq %in% cells_vdj]
```


### b. filter rna-seq and vdj data
* Even after filtering for cells that overlap with BCR/TCR data, there are still sparse clusters of cells annotated as different cell types
```{r}
# only keep cells with vdj
rnaseq_obj     <- rnaseq_obj_full[,colnames(rnaseq_obj_full) %in% overlapping_cells]

# add column to show which cells overlap with scRNA-Seq
bcr_df$cell_overlap_status <- "no"
tcr_df$cell_overlap_status <- "no"

bcr_df[bcr_df$unique_cell_id %in% overlapping_cells,"cell_overlap_status"] <- "yes"
tcr_df[tcr_df$unique_cell_id %in% overlapping_cells,"cell_overlap_status"] <- "yes"
```

# 4. Plot cell before and after VDJ filter

### a. SingleR annotations pre and post VDJ filter
```{r,fig.height=6,fig.width=14}
# generate UMAP of cell type annotations pre-filtering
p1_pre_filter_singler <- DimPlot(object = rnaseq_obj_full, group.by="main_bencode", pt.size=1.0, label.size = 5, alpha = 0.7,label = T,reduction = "umap")


# post filter umap
p2_post_filter_singler <- DimPlot(object = rnaseq_obj, group.by="main_bencode", pt.size=1.0, label.size = 5, alpha = 0.7,label = T,reduction = "umap")

plot_grid(p1_pre_filter_singler,p2_post_filter_singler,ncol = 2)
```

### b. Azimuth annotations pre and post VDJ filter
```{r,fig.height=6,fig.width=14}
# generate UMAP of cell type annotations pre-filtering
p1_pre_filter_azimuth <- DimPlot(object = rnaseq_obj_full, group.by="predicted.celltype.l1", pt.size=1.0, label.size = 5, alpha = 0.7,label = T,reduction = "umap")


# post filter umap
p2_post_filter_azimuth <- DimPlot(object = rnaseq_obj, group.by="predicted.celltype.l1", pt.size=1.0, label.size = 5, alpha = 0.7,label = T,reduction = "umap")

plot_grid(p1_pre_filter_azimuth,p2_post_filter_azimuth,ncol = 2)
```


# 5. Add columns to VDJ data
* This code block gets stuck when running. To fix this press **return** in the R console
```{r}
# Add sample column
tcr_df$sample <- tstrsplit(tcr_df$unique_cell_id,"-")[[2]]
bcr_df$sample <- tstrsplit(bcr_df$unique_cell_id,"-")[[2]]

# Add column for CSF/PB 
tcr_df$COMPARTMENT <- tstrsplit(tcr_df$sample,"_")[[2]]
bcr_df$COMPARTMENT <- tstrsplit(bcr_df$sample,"_")[[2]]
tcr_df$COMPARTMENT[tcr_df$COMPARTMENT %in% "PBMC"] <- "PB"
bcr_df$COMPARTMENT[bcr_df$COMPARTMENT %in% "PBMC"] <- "PB"

# Add Status Disease/Healthy
tcr_df$STATUS <- tstrsplit(tcr_df$sample,"_")[[1]] %>% as.character()
tcr_df$STATUS[tcr_df$STATUS %in% "32"] <- "Healthy"
tcr_df$STATUS[tcr_df$STATUS %in% "28"] <- "Disease"

bcr_df$STATUS <- tstrsplit(bcr_df$sample,"_")[[1]] %>% as.character()
bcr_df$STATUS[bcr_df$STATUS %in% "32"] <- "Healthy"
bcr_df$STATUS[bcr_df$STATUS %in% "28"] <- "Disease"
```


# 6. Re-cluster

* This line should only be run once
```{r,fig.height=6,fig.width=14}
set.seed(seed.number)
rnaseq_obj <- scaleAndClusterSeuratObject(rnaseq_obj,dims = 1:6,npca = 5,resolution = 0.3,dim.reduction = T)
```

# 7. Plot cell type annotations after re-clustering

### a. SingleR
```{r,fig.height=6,fig.width=14}
p3_reclustered_singler <- DimPlot(object = rnaseq_obj, group.by="main_bencode", pt.size=1.0, label.size = 5,label = T,reduction = "umap")
p4_reclustered_singler <- DimPlot(object = rnaseq_obj, group.by="fine_bencode", pt.size=1.0,label.size = 5, label = T,reduction = "umap")

plot_grid(p3_reclustered_singler,p4_reclustered_singler,ncol = 2)
```

### b. Azimuth
```{r,fig.height=6,fig.width=17}
p3_reclustered_azimuth <- DimPlot(object = rnaseq_obj, group.by="predicted.celltype.l1", pt.size=1.0, label.size = 5,label = T,reduction = "umap")
p4_reclustered_azimuth <- DimPlot(object = rnaseq_obj, group.by="predicted.celltype.l2", pt.size=1.0,label.size = 5, label = T,reduction = "umap")

plot_grid(p3_reclustered_azimuth,p4_reclustered_azimuth,ncol = 2)
```


### ---------------

# Manual cell type annotation

# 1. Look at the data

### a. Add columns and look at UMAP plots
```{r,fig.height=6,fig.width=17}
# Add compartment column
rnaseq_obj$compartment <- tstrsplit(rnaseq_obj$sample,"_")[[2]]
rnaseq_obj$compartment <- str_replace_all(rnaseq_obj$compartment,"PBMC","PB")

# Add patient status column
rnaseq_obj$status      <- tstrsplit(rnaseq_obj$sample,"_")[[1]]
rnaseq_obj$status[rnaseq_obj$status %in% "32"] <- "Healthy"
rnaseq_obj$status[rnaseq_obj$status %in% "28"] <- "Disease"

# loc 
loc_cols       = c("CSF" = "blue", "PB" = "red")
diagnosis_cols = c("Disease" = "purple","Healthy" = "orange")

p1 <- DimPlot(object = rnaseq_obj, group.by="seurat_clusters", pt.size=1.0, label.size = 10, alpha = 0.7,label = T,reduction = "umap")
p2 <- DimPlot(object = rnaseq_obj, group.by="compartment", pt.size=1.0, label.size = 10, alpha = 0.7,label = F,cols = loc_cols,reduction = "umap")
p3 <- DimPlot(object = rnaseq_obj, group.by="status", pt.size=1.0, label.size = 10,alpha = 0.7,label = F,cols = diagnosis_cols,reduction = "umap")

plot_grid(p1,p2,p3,ncol = 3)
```


### b. feature/violin of canonical genes
* For an extended list of gene markers for cell types, refer to **Celltype_Markers.xlsx** in the **resources** directory
* PPBP = Platelets, which are noise
```{r, fig.height=8, fig.width=10,warning=FALSE}
tcell_panel           <- c("CD3E","CD3D","CD3G","CD4","CD8A","CD8B")
bcell_panel           <- c("MS4A1","CD79A","CD79B","CD19","IGHM","IGHA1")
other_celltypes_panel <- c("CD14","LYZ","MS4A7","FCER1A","APOE","PPBP")
```


#### B cell gene panel
```{r, fig.height=8, fig.width=10,warning=FALSE}
FeaturePlot(rnaseq_obj,features = bcell_panel,ncol = 3)
```

#### T cell gene panel
```{r, fig.height=8, fig.width=10,warning=FALSE}
FeaturePlot(rnaseq_obj,features = tcell_panel,ncol = 3)
```

#### gene panel for other cell types

* We should not see an defined clusters for other cells types since these data
  should only contain B cells and T cells
* CD14+ Monocytes - LYZ, CD14
* CD16+ Monocytes - MS4A7
* Monocyte derived Dendritic Cells - FCGR3A,MS4A7
* Macrophage - APOE
* Platelet - PPBP
```{r, fig.height=8, fig.width=10,warning=FALSE}
FeaturePlot(rnaseq_obj,features = other_celltypes_panel,ncol = 3)
```

# 2. Add main cell types
* While cell types can be annotated manually based on gene expression, The 10X VDJ results tell us with higher confidence which cells are B cells/T cells

```{r,fig.height=6,fig.width= 12}
main_annot <- names(rnaseq_obj$orig.ident)
main_annot[main_annot %in% bcr_df$unique_cell_id] <- "B cells"
main_annot[main_annot %in% tcr_df$unique_cell_id] <- "T cells"
names(main_annot) <- names(rnaseq_obj$orig.ident)
rnaseq_obj$main.celltype.annot <- main_annot

p4 <- DimPlot(object = rnaseq_obj, group.by="seurat_clusters", pt.size=1.0, label.size = 5, alpha = 0.7,label = T,reduction = "umap")
p5 <- DimPlot(object = rnaseq_obj, group.by="main.celltype.annot", pt.size=1.0, label.size = 5, alpha = 0.7,label = T,reduction = "umap")

plot_grid(p4,p5, ncol = 2)
```

# 3. Add in Subtype annotations

### a. Violin of CD4/CD8 genes
```{r,fig.height=8,fig.width= 12}
subtype_annotations <- colnames(rnaseq_obj)

Idents(rnaseq_obj) <- rnaseq_obj$seurat_clusters
p6 <- VlnPlot(rnaseq_obj,features = c("CD4","CD8A","CD8B"),pt.size = 0,ncol = 1)

plot_grid(p6,p4,ncol = 2,rel_heights = c(8,6),rel_widths = c(6,6))
```

### b. Define T cell subtypes based on CD8 gene expression & SingleR annotations
* CD8 T cells are much higher in automated SingleR annotations
* In addition, there are cell types which are mis-annotated as Macrophages, Monocytes and NK cells
* Mis-annotated celltype cluster with B cells
* We are going to re-annotate T cells manually based on CD8 expression
```{r,fig.height=6,fig.width=17}
FeaturePlot(rnaseq_obj, features = c("CD8A", "CD8B"),blend = TRUE)
```

```{r,fig.height=9,fig.width= 10}
# SingleR and Azimuth annotatons
celltype_annot_df <- rnaseq_obj@meta.data[,c("sample","seurat_clusters","main_bencode","fine_bencode","main.celltype.annot","predicted.celltype.l1","predicted.celltype.l2")]

# Manual T cell annotation based on CD8 expression
CD8_pos_tcells           <- colnames(subset(rnaseq_obj, subset = CD8A > 0 | CD8B > 0))
celltype_annot_df$cd8_pos <- row.names(celltype_annot_df)
celltype_annot_df$cd8_pos[celltype_annot_df$cd8_pos %in% CD8_pos_tcells] <- "yes"
celltype_annot_df$cd8_pos[! celltype_annot_df$cd8_pos %in% "yes"]        <- "no"

tcell_subtype_annotation <- names(rnaseq_obj$main.celltype.annot[rnaseq_obj$main.celltype.annot %in% "T cells"])
CD8_pos_tcells <- tcell_subtype_annotation[tcell_subtype_annotation %in% CD8_pos_tcells]
CD4_tcells     <- tcell_subtype_annotation[! tcell_subtype_annotation %in% CD8_pos_tcells]

celltype_annot_df$sub.celltype.annot <- celltype_annot_df$main.celltype.annot
celltype_annot_df[row.names(celltype_annot_df) %in% CD8_pos_tcells, "sub.celltype.annot"] <- "CD8 T cells"
celltype_annot_df[row.names(celltype_annot_df) %in% CD4_tcells, "sub.celltype.annot"]     <- "CD4 T cells"

# Stacked bar
# - Not all Celltype predictions for CD8+ T cells express CD8

# SingleR
summary_main_bencode <- celltype_annot_df %>% count(seurat_clusters, main_bencode, name = "count")
levels(summary_main_bencode$main_bencode) <- sort(unique(summary_main_bencode$main_bencode))
colors_main_bencode <- c("red","darkgreen","darkblue","purple","orange","grey")
p7 <- ggplot(summary_main_bencode, aes(x = seurat_clusters, y = count, fill = main_bencode)) + geom_bar(stat = "identity",position = "fill") + scale_fill_manual(values = colors_main_bencode) + ggtitle("SingleR Cell-type Predictions")

# Azimuth
summary_azimuth <- celltype_annot_df %>% count(seurat_clusters, predicted.celltype.l1, name = "count")
levels(summary_azimuth$predicted.celltype.l1) <- sort(unique(summary_azimuth$predicted.celltype.l1))
colors_azimuth <- c("red","darkgreen","darkblue","purple","orange","brown","black","grey")
p8 <- ggplot(summary_azimuth, aes(x = seurat_clusters, y = count, fill = predicted.celltype.l1)) + geom_bar(stat = "identity",position = "fill") + scale_fill_manual(values = colors_azimuth) + ggtitle("Azimuth Cell-type Predictions")

# Annotations based on BCR and TCR VDJ data
summary_subtype <- celltype_annot_df %>% count(seurat_clusters, sub.celltype.annot, name = "count")
levels(summary_subtype$sub.celltype.annot) <- sort(unique(summary_subtype$sub.celltype.annot))
colors_subtype <- c("red","darkgreen","darkblue")
p9 <- ggplot(summary_subtype, aes(x = seurat_clusters, y = count, fill = sub.celltype.annot)) + geom_bar(stat = "identity",position = "fill") + scale_fill_manual(values = colors_subtype) + ggtitle("Multiomics")

plot_grid(p7,p8,p9,ncol = 1,align = "v")
```

### c. Define B cell subtypes based on Azimuth subtype annotations

```{r, fig.height=4, fig.width=10,warning=FALSE}
celltype_annot_df_bcr <- celltype_annot_df[celltype_annot_df$main.celltype.annot %in% "B cells",]
uni_bcell_subtypes   <- unique(celltype_annot_df$predicted.celltype.l2)
non_bcell_subtypes   <- uni_bcell_subtypes[! uni_bcell_subtypes %in% c("B naive","B memory","B intermediate","Plasmablast")]

# For simplicity we will annotate non-B cell annotations as Naive B cells
#  - Violin plots showed that these cells have virtually no IgG, IgA or V gene expression
celltype_annot_df_bcr$predicted.celltype.l2[celltype_annot_df_bcr$predicted.celltype.l2 %in% non_bcell_subtypes] <- "B naive"

# Add B cell subtype annotations into sub.celltype.annot column
for (bsubtype in unique(celltype_annot_df_bcr$predicted.celltype.l2)) {
  subset_cells <- row.names(celltype_annot_df_bcr[celltype_annot_df_bcr$predicted.celltype.l2 %in% bsubtype,])
  celltype_annot_df[row.names(celltype_annot_df) %in% subset_cells,"sub.celltype.annot"] <- bsubtype
}


# Add subtype annotations back into RNA-Seq object
rnaseq_obj$sub.celltype.annot <- celltype_annot_df$sub.celltype.annot

# Plot
p10  <- dittoDimPlot(rnaseq_obj, "main.celltype.annot")
p11 <- dittoDimPlot(rnaseq_obj, "sub.celltype.annot")

grid.arrange(p10,p11,nrow=1)
```


# 4. Add cell type annotation to VDJ results
```{r}
# TCR 
tcell_annotation_col        <- rep("No Annot",nrow(tcr_df))
names(tcell_annotation_col) <- tcr_df$unique_cell_id

subtype_annotations_tcr <- singler_annot_df[singler_annot_df$main.celltype.annot %in% "T cells",]
for (subtype in unique(subtype_annotations_tcr$sub.celltype.annot)) {
  subtype_cells <- row.names(subtype_annotations_tcr[subtype_annotations_tcr$sub.celltype.annot %in% subtype,])
  tcell_annotation_col[names(tcell_annotation_col) %in% subtype_cells] <- subtype
}
tcr_df$SUBTYPE_ANNOTATION <- tcell_annotation_col

# BCR
bcell_annotation_col<- rep("No Annot",nrow(bcr_df))
names(bcell_annotation_col) <- bcr_df$unique_cell_id

subtype_annotations_bcr <- singler_annot_df[singler_annot_df$main.celltype.annot %in% "B cells",]
for (subtype in unique(subtype_annotations_bcr$sub.celltype.annot)) {
  subtype_cells    <- row.names(subtype_annotations_bcr[subtype_annotations_bcr$sub.celltype.annot %in% subtype,])
  bcell_annotation_col[names(bcell_annotation_col) %in% subtype_cells] <- subtype
}
bcr_df$SUBTYPE_ANNOTATION <- bcell_annotation_col
```


# 5. define expanded clonotypes
* createCloneCountTable & addExpansionCategories are in functions_workshop.R

### a. generate clone count data frames

* The clone counts are calculated based on the full VDJ data
```{r}
clone_counts_tcr <- createCloneCountTable(tcr_df)
clone_counts_bcr <- createCloneCountTable(bcr_df)

# We can identify expanded clonotypes in the CSF with a high CSF:PB ratio
clone_count_subset <- clone_counts_tcr[clone_counts_tcr$cell_count > 1 & clone_counts_tcr$CSF_count > 1, ]
```


### b. create categories for clonal expansion
* Highly expanded - cell count > 5
* Moderate        - 1 < cells count < 5
* Non-expanded    - cell count = 1
```{r}
tcr_df <- addExpansionCategories(tcr_df,clone_counts_tcr)
bcr_df <- addExpansionCategories(bcr_df,clone_counts_bcr)
```

### c. add expansion categories to scRNA-Seq object
```{r}
expansion_categ_col <- paste(colnames(rnaseq_obj),rnaseq_obj$main.celltype.annot)
uni_expansion_categ <- unique(tcr_df$EXPANSION_CATEGORY)

for (categ in unique(tcr_df$EXPANSION_CATEGORY)) {
  cells_categ <- c()
  cells_tcr <- tcr_df[tcr_df$EXPANSION_CATEGORY %in% categ,"unique_cell_id"]
  cells_tcr <- unique(paste(cells_tcr,"T cells"))
  cells_bcr <- bcr_df[bcr_df$EXPANSION_CATEGORY %in% categ,"unique_cell_id"]
  cells_bcr <- unique(paste(cells_bcr,"B cells"))
  
  if(length(cells_tcr)> 0) {cells_categ <- c(cells_categ, cells_tcr)}
  if(length(cells_bcr)> 0) {cells_categ <- c(cells_categ, cells_bcr)}
  
  expansion_categ_col[expansion_categ_col %in% cells_categ] <- categ
}
names(expansion_categ_col) <- colnames(rnaseq_obj)
rnaseq_obj$expansion_category <- expansion_categ_col
```


# 6. Look at proportions

### a. subset B cells
```{r}
# add column for Diagnosis + Subtype & Diagnois + Compartment
rnaseq_obj$status.subtype     <- paste(rnaseq_obj$status,rnaseq_obj$sub.celltype.annot)
rnaseq_obj$status.compartment <- paste(rnaseq_obj$status,rnaseq_obj$compartment)

cells <- names(rnaseq_obj$main.celltype.annot[rnaseq_obj$main.celltype.annot %in% "B cells"])
bcell_obj <- rnaseq_obj[,cells]
cells <- names(rnaseq_obj$main.celltype.annot[rnaseq_obj$main.celltype.annot %in% "T cells"])
tcell_obj <- rnaseq_obj[,cells]
```

### *b. sample breakdown
* Disease Patient has an increased proportion of B cells in the CSF of which most are memory B cells
```{r,fig.height=6, fig.width=7,warning=FALSE}
p12 <- dittoBarPlot(rnaseq_obj, "sub.celltype.annot", group.by = "status.compartment", main = "All cells",color.panel = dittoColors()[1:6])
p13 <- dittoBarPlot(bcell_obj, "sub.celltype.annot", group.by = "status.compartment", main = "B cells",color.panel = dittoColors()[1:4])


plot_grid(p12,p13,nrow=2,align = "v")
```

### c. clonotype expansion breakdown
* Expanded T cell populations are observed in all compartments, while expanded B cells show more restriction
```{r,fig.height=6, fig.width=7,warning=FALSE}
p14 <- dittoBarPlot(tcell_obj, "expansion_category", group.by = "status.compartment", main = "T cells")
p15 <- dittoBarPlot(bcell_obj, "expansion_category", group.by = "status.compartment", main = "B cells")

grid.arrange(p14,p15,nrow=2)
```


### ---------------

# SAVE Processed objects
```{r,eval=FALSE}
save(rnaseq_obj,tcell_obj,bcell_obj,bcr_df,tcr_df,clone_counts_tcr,clone_counts_bcr,file = "objects/processed_objects.RData")
```

