---
title: "Filter scRNA-Seq and VDJ data"
author: "Ravi Dandekar"
date: "4/19/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

GOAL: annotate cell types and define expanded clonotypes (B cells and T cells)

* Before starting set working directory to the CompImmunologyWorkshopUCSF folder
  * setwd("[PATH]/CompImmunologyWorkshopUCSF")

```{r,echo=FALSE,warning=FALSE}
# Set Working Directory before starting

# Immcantation
library(alakazam)
library(shazam)
library(tigger)
library(dowser)

library(Seurat)
library(dittoSeq)
library(reshape2)
library(kableExtra)
library(devtools)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(gridExtra)    # grid.arrange()
library(cowplot)
library(dplyr)
library(impactSingleCellToolkit)

# Immcantation Results
bcr_heavy_fh <- "data/results_immcantation/BCR_CSFPB_heavy_germ-pass.tsv"
bcr_light_fh <- "data/results_immcantation/BCR_CSFPB_light_germ-pass.tsv"
tcr_beta_fh  <- "data/results_immcantation/TCR_CSFPB_heavy_germ-pass.tsv"
tcr_alpha_fh <- "data/results_immcantation/TCR_CSFPB_light_germ-pass.tsv"

source("resources/functions_workshop.R")
```


# 1. Load RNA-Seq object

At this stage of the analysis scRNA-Seq, BCR and TCR repertoire datasets have been processed and overlapped
* Doublet were removed computationally using 'DoubletFinder'
* cells with more than 10% mitochondrial gene expression were omitted
* Only B cells and T cells exist in this object

```{r}
rnaseq_obj <- readRDS("objects/seurat.obj.processed.rds")
```

# 2. Load and format TCR/BCR repertoire results

* Clonotype definitions for BCR are based on the Heavy chain only
* Clonotype definitions for TCR include both Alpha and Beta chains

### a. Load immcantations results
```{r}
# TCR
db.tra   <- readChangeoDb(tcr_alpha_fh) %>% as.data.frame()
db.trb   <- readChangeoDb(tcr_beta_fh) %>% as.data.frame()
# BCR
db.heavy <- readChangeoDb(bcr_heavy_fh) %>% as.data.frame()
db.light <- readChangeoDb(bcr_light_fh) %>% as.data.frame()
```

### b. filter contigs and merge all chains
* All contigs are in frame
```{r}
bcr_df <- mergeVDJresults(db.heavy,db.light,umi.thresh = 3,assay = "bcr")
tcr_df <- mergeVDJresults(db.trb,db.tra,umi.thresh = 2, assay = "tcr")

# format VDJ unique cell IDs to match scRNA-Seq (function only for this workshop)
formatUniqueIDcolum <- function(uni_id_column,sep = ":") {
  samples <- tstrsplit(uni_id_column, sep)[[1]]
  cells   <- tstrsplit(uni_id_column, sep)[[2]]
  fmt_ids <- paste(cells,samples,sep = "-")
  
  return(fmt_ids)
}

bcr_df$unique_cell_id <- formatUniqueIDcolum(bcr_df$unique_cell_id)
tcr_df$unique_cell_id <- formatUniqueIDcolum(tcr_df$unique_cell_id)
```


# 3. Only analyze cells shared across all datasets

### a. get vector of overlapping cells
```{r}
cells_rnaseq      <- colnames(rnaseq_obj)
cells_vdj         <- unique(c(bcr_df$unique_cell_id,
                              tcr_df$unique_cell_id))
overlapping_cells <- cells_rnaseq[cells_rnaseq %in% cells_vdj]
```

### b. filter rnaseq and vdj data
```{r,fig.height=6,fig.width=11}
# generate UMAP of cell type annotations pre-filtering
p1_pre_filter <- UMAPPlot(object = rnaseq_obj, group.by="main_bencode", pt.size=0.3, label = T)

# only keep cells with vdj
rnaseq_obj     <- rnaseq_obj[,colnames(rnaseq_obj) %in% overlapping_cells]
bcr_df         <- bcr_df[bcr_df$unique_cell_id %in% overlapping_cells,]
tcr_df         <- tcr_df[tcr_df$unique_cell_id %in% overlapping_cells,]

# post filter umap
p2_post_filter <- UMAPPlot(object = rnaseq_obj, group.by="main_bencode", pt.size=0.3, label = T)
plot_grid(p1_pre_filter,p2_post_filter,ncol = 2)
```

# 4. Re-cluster
```{r,fig.height=6,fig.width=14}
rnaseq_obj <- scaleAndClusterSeuratObject(rnaseq_obj,dims = 1:5,npca = 5,resolution = 0.3,
                                          normalize = F,dim.reduction = T)

p3_reclustered <- UMAPPlot(object = rnaseq_obj, group.by="main_bencode", pt.size=0.3, label = T)
p4_reclustered <- UMAPPlot(object = rnaseq_obj, group.by="fine_bencode", pt.size=0.3, label = T)

plot_grid(p3_reclustered,p4_reclustered,ncol = 2)
```

# ---------------

# Manual Cell type annotation

# 1. Look at the data

### a. Add columns and look at UMAP plots
```{r,fig.height=6,fig.width=17}
# Add compartment column
rnaseq_obj$compartment <- tstrsplit(rnaseq_obj$sample,"_")[[2]]
rnaseq_obj$compartment <- str_replace_all(rnaseq_obj$compartment,"PBMC","PB")

# Add patient status column
rnaseq_obj$status      <- tstrsplit(rnaseq_obj$sample,"_")[[1]]
rnaseq_obj$status[rnaseq_obj$status %in% "32"] <- "Healthy"
rnaseq_obj$status[rnaseq_obj$status %in% "28"] <- "Disease"

# loc 
loc_cols       = c("CSF" = "blue", "PB" = "red")
diagnosis_cols = c("Disease" = "purple","Healthy" = "orange")

p1 <- UMAPPlot(object = rnaseq_obj, group.by="seurat_clusters", pt.size=0.3, label = T)
p2 <- UMAPPlot(object = rnaseq_obj, group.by="compartment", pt.size=0.3, label = F,cols = loc_cols)
p3 <- UMAPPlot(object = rnaseq_obj, group.by="status", pt.size=0.3, label = F,cols = diagnosis_cols)

plot_grid(p1,p2,p3,ncol = 3)
```


### b. feature/violin of canonical genes

* PPBP = Platelets, which are noise
```{r, fig.height=8, fig.width=10,echo=FALSE,warning=FALSE}
tcell_panel           <- c("CD3E","CD3D","CD3G","CD4","CD8A","CD8B")
bcell_panel           <- c("MS4A1","CD79A","CD79B","CD19","IGHM","IGHA1")
other_celltypes_panel <- c("CD14","LYZ","MS4A7","FCER1A","APOE","PPBP")
```


#### B cell gene panel
```{r, fig.height=8, fig.width=10,echo=FALSE,warning=FALSE}
FeaturePlot(rnaseq_obj,features = bcell_panel,ncol = 3)
```

#### T cell gene panel
```{r, fig.height=8, fig.width=10,echo=FALSE,warning=FALSE}
FeaturePlot(rnaseq_obj,features = tcell_panel,ncol = 3)
```

#### gene panel for other cell types

* Should not see an defined clusters for other cells types since these data
  should only contain B cells and T cells
```{r, fig.height=8, fig.width=10,echo=FALSE,warning=FALSE}
FeaturePlot(rnaseq_obj,features = other_celltypes_panel,ncol = 3)
```

# 2. Add main cell types
* While cell types can be annotated manually based on gene expression, The 10X VDJ results tell us with higher confidence which cells are B cells/T cells

```{r,fig.height=6,fig.width= 12}
main_annot <- names(rnaseq_obj$orig.ident)
main_annot[main_annot %in% bcr_df$unique_cell_id] <- "B cells"
main_annot[main_annot %in% tcr_df$unique_cell_id] <- "T cells"
names(main_annot) <- names(rnaseq_obj$orig.ident)
rnaseq_obj$main.celltype.annot <- main_annot

p4 <- UMAPPlot(object = rnaseq_obj, group.by="seurat_clusters", pt.size=0.3, label = T)
p5 <- UMAPPlot(object = rnaseq_obj, group.by="main.celltype.annot", pt.size=0.3, label = T)

plot_grid(p4,p5, ncol = 2)
```

# 3. Add in Subtype annotations

### a. Violin of CD4/CD8 genes
```{r,fig.height=8,fig.width= 12}
subtype_annotations <- colnames(rnaseq_obj)

Idents(rnaseq_obj) <- rnaseq_obj$seurat_clusters
p6 <- VlnPlot(rnaseq_obj,features = c("CD4","CD8A","CD8B"),pt.size = 0,ncol = 1)

plot_grid(p6,p4,ncol = 2,rel_heights = c(8,6),rel_widths = c(6,6))
```

### b. Define T cell subtypes based on CD8 gene expression & SingleR annotations

#### --colors
```{r}
colors_main_bencode <- c("darkblue","darkgreen","purple","red","orange","grey")
colors_main_bencode       = c("CD4+ T-cells" = "blue", 
                              "CD8+ T-cells" = "darkgreen",
                              "NK cells" = "purple",
                              "B-cells" = "red",
                              "Monocytes" = "orange",
                              "Macrophages" = "darkgrey")
diagnosis_cols = c("CD4 T cells" = "blue",
                   "CD8 T cells" = "darkgreen",
                   "B cells" = "red")
```

```{r,fig.height=6,fig.width= 10}
# SingleR Encode annotatons
singler_annot_df <- rnaseq_obj@meta.data[,c("sample","seurat_clusters","main_bencode","fine_bencode","main.celltype.annot")]

# Manual T cell annotation based on CD8 expression
CD8_pos_tcells           <- colnames(subset(rnaseq_obj, subset = CD8A > 0 | CD8B > 0))
singler_annot_df$cd8_pos <- row.names(singler_annot_df)
singler_annot_df$cd8_pos[singler_annot_df$cd8_pos %in% CD8_pos_tcells] <- "yes"
singler_annot_df$cd8_pos[! singler_annot_df$cd8_pos %in% "yes"]        <- "no"

tcell_subtype_annotation <- names(rnaseq_obj$main.celltype.annot[rnaseq_obj$main.celltype.annot %in% "T cells"])
CD8_pos_tcells <- tcell_subtype_annotation[tcell_subtype_annotation %in% CD8_pos_tcells]
CD4_tcells     <- tcell_subtype_annotation[! tcell_subtype_annotation %in% CD8_pos_tcells]

singler_annot_df$sub.celltype.annot <- singler_annot_df$main.celltype.annot
singler_annot_df[row.names(singler_annot_df) %in% CD8_pos_tcells, "sub.celltype.annot"] <- "CD8 T cells"
singler_annot_df[row.names(singler_annot_df) %in% CD4_tcells, "sub.celltype.annot"]     <- "CD4 T cells"

# Stacked bar
# - Not all SingleR predictions for CD8+ T cells express CD8
summary_main_bencode <- singler_annot_df %>% count(seurat_clusters, main_bencode, name = "count")
levels(summary_main_bencode$main_bencode) <- sort(unique(summary_main_bencode$main_bencode))
colors_main_bencode <- c("red","darkgreen","darkblue","purple","orange","grey")
p7 <- ggplot(summary_main_bencode, aes(x = seurat_clusters, y = count, fill = main_bencode)) + geom_bar(stat = "identity",position = "fill") + scale_fill_manual(values = colors_main_bencode) + ggtitle("SingleR Cell-type Predictions")

summary_subtype <- singler_annot_df %>% count(seurat_clusters, sub.celltype.annot, name = "count")
levels(summary_subtype$sub.celltype.annot) <- sort(unique(summary_subtype$sub.celltype.annot))
colors_subtype <- c("red","darkgreen","darkblue")
p8 <- ggplot(summary_subtype, aes(x = seurat_clusters, y = count, fill = sub.celltype.annot)) + geom_bar(stat = "identity",position = "fill") + scale_fill_manual(values = colors_subtype) + ggtitle("Multiomics")

plot_grid(p7,p8,ncol = 1,align = "v")
```

### c. Define B cell subtypes based on SingleR automated annotations

* SingleR
* Training Data ENCODE

```{r, fig.height=4, fig.width=10,echo=FALSE,warning=FALSE}
singler_annotations <- read.csv("resources/Bcell_subtype_annotations.SingleR.csv")

for (subtype in unique(singler_annotations$subtype)) {
  subtype_df    <- singler_annotations[singler_annotations$subtype %in% subtype,]
  subtype_annotations[subtype_annotations %in% subtype_df$cell_barcode] <- subtype
}
rnaseq_obj$sub.celltype.annot <- subtype_annotations

# Plot
p7 <- dittoDimPlot(rnaseq_obj, "main.celltype.annot")
p8 <- dittoDimPlot(rnaseq_obj, "sub.celltype.annot")

grid.arrange(p7,p8,nrow=1)
```


# 4. Add cell type annotation to VDJ results
```{r}
# TCR 
tcell_annotation_col<- rep("No Annot",length(tcr_results$CELL_BARCODES))
names(tcell_annotation_col) <- tcr_results$CELL_BARCODES

tcell_annotation_col[names(tcell_annotation_col) %in% CD4_tcells]     <- "CD4 T cells"
tcell_annotation_col[names(tcell_annotation_col) %in% CD8_pos_tcells] <- "CD8 T cells"
tcr_results$SUBTYPE_ANNOTATION                                        <- tcell_annotation_col

# BCR
bcell_annotation_col<- rep("No Annot",length(bcr_results$CELL_BARCODES))
names(bcell_annotation_col) <- bcr_results$CELL_BARCODES

for (subtype in unique(singler_annotations$subtype)) {
  subtype_df    <- singler_annotations[singler_annotations$subtype %in% subtype,]
  bcell_annotation_col[names(bcell_annotation_col) %in% subtype_df$cell_barcode] <- subtype
}
bcr_results$SUBTYPE_ANNOTATION <- bcell_annotation_col
```


# ---------------

# SAVE Processed objects
```{r}
# 2022 command
#save(rnaseq_obj,tcell_obj,bcell_obj,bcr_results,tcr_results,clone_counts_tcr,clone_counts_bcr,file = "objects/processed_objects.RData")
save(rnaseq_obj,bcr_df,tcr_df,file = "objects/processed_objects.RData")
```

# SAVE ENV - Temporary
```{r}
save(list = ls(),file = "objects/TEMP_ENV.RData")
```