---
title: "RNA-Seq Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


* Before starting set working directory to the CompImmunologyWorkshopUCSF folder
  * setwd("[PATH]/CompImmunologyWorkshopUCSF")

```{r,warning=FALSE}
# Set Working Directory before starting

# Immcantation
library(alakazam)
library(shazam)
library(tigger)

library(Seurat)
library(dittoSeq)
library(reshape2)
library(kableExtra)
library(devtools)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(gridExtra)    # grid.arrange()
library(cowplot)
library(dplyr)
library(impactSingleCellToolkit)

source("resources/functions_workshop.R")
```


# 1. Load Processed Objects
```{r}
load("objects/processed_objects.RData")
```


# 2. Subset Disease Objects

### a. subset B and T cells
```{r}
# T cell CSF object
tcell_obj_csf <- tcell_obj[,names(tcell_obj$status[tcell_obj$status %in% "Disease"])]
tcell_obj_csf <- tcell_obj_csf[,names(tcell_obj_csf$compartment[tcell_obj_csf$compartment %in% "CSF"])]
tcell_obj_csf$expansion_category[tcell_obj_csf$expansion_category %in% c("High","Moderate")] <- "Expanded"
tcell_obj_csf$expansion.subtype <- paste(tcell_obj_csf$expansion_category,tcell_obj_csf$sub.celltype.annot)


# B cell object
bcell_obj_csfpb <- bcell_obj
bcell_obj_csfpb$expansion_category[bcell_obj_csfpb$expansion_category %in% c("High","Moderate")] <- "Expanded"
bcell_obj_csfpb$expansion.subtype <- paste(bcell_obj_csfpb$expansion_category,bcell_obj_csfpb$sub.celltype.annot)
```

### b. B cell gene usage (heatmap)

* With RNA-Seq alone, it is hard to separate out clonotypes

#### V genes
```{r,fig.height=7,fig.width=10}
genes_bcells             <- row.names(bcell_obj_csfpb)
vgenes_bcells            <- sort(genes_bcells[grep("IGHV",genes_bcells)])

h1 <- dittoHeatmap(bcell_obj_csfpb, genes = vgenes_bcells,
        annot.by = c("sub.celltype.annot","compartment","status"),
        scaled.to.max = TRUE,
        show_colnames = FALSE,
        show_rownames = TRUE,
        cluster_cols = TRUE,
        main = "V gene usage B cells")

```

#### Other Ig genes
```{r,fig.height=5,fig.width=10}
genes_bcells             <- row.names(bcell_obj_csfpb)
other_ig_genes_bcells    <- sort(genes_bcells[grep("IGHA|IGHG|IGHM|IGHD",genes_bcells)])

h2 <- dittoHeatmap(bcell_obj_csfpb, genes = other_ig_genes_bcells,
        annot.by = c("sub.celltype.annot","compartment","status"),
        scaled.to.max = TRUE,
        show_colnames = FALSE,
        show_rownames = TRUE,
        cluster_cols = FALSE,
        main = "Ig gene usage B cells")

```



# 3. Differential Gene Expression

* If it slows down, pre-load DGE results for CSF vs. PB
```{r}
load("objects/csf_vs_pb_DGE.RData")
```

### a. CSF vs. PB
```{r}
TOP_N <- 100

# Tcells
Idents(tcell_obj) <- tcell_obj$compartment
levels(tcell_obj) <- c("CSF","PB")

dge_compartment_tcells_all <- FindAllMarkers(tcell_obj,logfc.threshold = 0.0,test.use = "wilcox")
dge_compartment_tcells <- dge_compartment_tcells_all[dge_compartment_tcells_all$p_val_adj < 0.05,]
top_genes_tcells       <- dge_compartment_tcells %>% group_by(cluster) %>% top_n(TOP_N, avg_log2FC)

# B cells
Idents(bcell_obj) <- bcell_obj$compartment
levels(bcell_obj) <- c("CSF","PB")

dge_compartment_bcells_all <- FindAllMarkers(bcell_obj,logfc.threshold = 0.0,test.use = "wilcox")
dge_compartment_bcells <- dge_compartment_bcells_all[dge_compartment_bcells_all$p_val_adj < 0.05,]
top_genes_bcells       <- dge_compartment_bcells %>% group_by(cluster) %>% top_n(TOP_N, avg_log2FC)
```

#### Heatmaps

```{r,fig.height=6,fig.width=9}
p1 <- dittoHeatmap(tcell_obj, genes = top_genes_tcells$gene,
        annot.by = c("compartment","status", "sub.celltype.annot"),
        scaled.to.max = TRUE,
        show_colnames = FALSE,
        show_rownames = FALSE,
        cluster_cols = FALSE,
        main = "CSF vs. PB T cells")
```

```{r,fig.height=6,fig.width=9}
p2 <- dittoHeatmap(bcell_obj, genes = top_genes_bcells$gene,
        annot.by = c("compartment","status", "sub.celltype.annot"),
        scaled.to.max = TRUE,
        show_colnames = FALSE,
        show_rownames = FALSE,
        cluster_cols = FALSE,
        main = "CSF vs. PB B cells")
```

#### Volcano Plots

```{r,fig.height=5,fig.width=9,warning=FALSE}
v1 <- plotDGEvolcano(dge_compartment_bcells_all,YLIM = c(0,50),title = "B cell DGE CSF vs. PB",cluster_filt = "CSF",
                    p.adj.thresh = 0.05,show_top_genes = 20)
v2 <- plotDGEvolcano(dge_compartment_tcells_all,YLIM = c(0,250),title = "T cell DGE CSF vs. PB",cluster_filt = "CSF",
                    p.adj.thresh = 0.05,show_top_genes = 20)

plot_grid(v1,v2,nrow = 1)
```


### b. T cells Expanded vs. Non-expanded (CSF)
```{r,fig.height=7, fig.width=8,echo=TRUE,warning=FALSE}
TOP_N <- 12
Idents(tcell_obj_csf) <- tcell_obj_csf$expansion.subtype
levels(tcell_obj_csf) <- c("Expanded CD8 T cells","Non-expanded CD8 T cells","Expanded CD4 T cells","Non-expanded CD4 T cells")

dge_tcell_csf   <- FindAllMarkers(tcell_obj_csf,logfc.threshold = 0.1,test.use = "wilcox")
dge_tcell_csf   <- dge_tcell_csf[dge_tcell_csf$p_val_adj < 0.05,]
top_genes_tcells_csf <- dge_tcell_csf %>% group_by(cluster) %>% top_n(TOP_N, avg_log2FC)

DotPlot(object = tcell_obj_csf,features = unique(top_genes_tcells_csf$gene)) + theme(axis.text.x = element_text(angle = 90, hjust=1)) + ggtitle("T cells CSF High vs. Non-Expanded") + coord_flip()
```




### c. B cells Expanded vs. Non-expanded (CSF and PB)
```{r,fig.height=10, fig.width=8,echo=TRUE,warning=FALSE}
TOP_N <- 10
Idents(bcell_obj_csfpb) <- bcell_obj_csfpb$expansion.subtype
complete_order          <- c("Expanded Plasma cells","Non-expanded Plasma cells",
                             "Expanded Memory B-cells","Non-expanded Memory B-cells",
                             "Expanded naive B-cells","Non-expanded naive B-cells")
complete_order_sub <- complete_order[complete_order %in% unique(bcell_obj_csfpb$expansion.subtype)]
levels(bcell_obj_csfpb) <- complete_order_sub

dge_bcell_csfpb   <- FindAllMarkers(bcell_obj_csfpb,logfc.threshold = 0.1,test.use = "wilcox")
dge_bcell_csfpb   <- dge_bcell_csfpb[dge_bcell_csfpb$p_val_adj < 0.05,]
top_genes_bcells_csfpb <- dge_bcell_csfpb %>% group_by(cluster) %>% top_n(TOP_N, avg_log2FC)

DotPlot(object = bcell_obj_csfpb,features = unique(top_genes_bcells_csfpb$gene)) + theme(axis.text.x = element_text(angle = 90, hjust=1)) + ggtitle("B cells CSF High vs. Non-Expanded") + coord_flip()
```

# 4. Pathway analysis

```{r,echo=FALSE,warning=FALSE}
library(ReactomePA)
library(org.Hs.eg.db)
library(igraph)
library(enrichplot)
library(ggnewscale)

library(DOSE)
library(clusterProfiler)
library(ggupset)
```

### a. B cells Memory
```{r,fig.height=6, fig.width=11,echo=TRUE,warning=FALSE}
#target_genes <- dge_bcell_csfpb[dge_bcell_csfpb$cluster %in% c("Expanded Memory B-cells","Non-expanded Memory B-cells"),]
target_genes <- dge_bcell_csfpb[dge_bcell_csfpb$cluster %in% c("Expanded Memory B-cells"),]

# omiit Ribosomal genes (RP)
#rp_genes     <- target_genes$gene[grep("^RP",target_genes$gene)]
#target_genes <- target_genes[! target_genes$gene %in% rp_genes,]

# Most of the expression should be in the target cluster
# - pct.1 = proportion of expression of the gene within the target cluster group
# - pct.2 = proportion of expression of the gene within everything outside the target cluster
#target_genes <- target_genes[target_genes$pct.1 > 0.5,]

# Convert gene names to Entrez IDs
de_genes <- mapIds(org.Hs.eg.db, target_genes$gene, 'ENTREZID', 'SYMBOL')
de_genes <- de_genes[!is.na(de_genes)]
de <- as.character(de_genes)

# Create FC and P-adj-value vectors for enrihed genes
de_fc_df <- target_genes[target_genes$gene %in% names(de_genes),]
entrez_col <- c()
for (g in names(de_genes)) {
  egene <- as.character(de_genes[g])
  entrez_col <- c(entrez_col,egene)
}
de_fc_df$entrez_id <- entrez_col

de_fc_list        <- de_fc_df$avg_log2FC
names(de_fc_list) <- de_fc_df$entrez_id

# P-value vec
de_pval_list        <- de_fc_df$p_val_adj
names(de_pval_list) <- de_fc_df$entrez_id

# Pathway Enrichment
bcell_pathways       <- enrichPathway(gene=de,pvalueCutoff=0.05, readable=T)
bcell_pathways_df    <- as.data.frame(bcell_pathways)

# Top enriched pathways and the gene count within each pathway
barplot(bcell_pathways, showCategory = 10)
```

#### Network of genes shared between pathways
```{r,fig.height=6, fig.width=15,echo=TRUE,warning=FALSE}
bcell_pathways_net <- pairwise_termsim(bcell_pathways,showCategory = 14)
emapplot(bcell_pathways_net, showCategory = 14,
         repel = T,cex_label_group = 0.5,layout = "star")
```



#### Network of Pathways and genes
```{r,fig.height=6,fig.width=15}
edo <- enrichDGN(as.list(de))
de_fc_list <- sort(de_fc_list,decreasing = T)

## convert gene ID to Symbol
edox <- setReadable(edo, 'org.Hs.eg.db', 'ENTREZID')

## Network categorySize can be scaled by 'pvalue' or 'geneNum' 
p1 <- cnetplot(edox, categorySize="pvalue", foldChange=de_fc_list)

## Heatmap
p2 <- heatplot(edox, foldChange=de_fc_list, showCategory=9)

plot_grid(p1,p2,ncol = 2)
```

#### Upset plot
```{r,fig.height=5,fig.width=7}
upset_plot <-upsetplot(edo)
upset_plot
```


### b. T cells CD8
```{r,fig.height=5, fig.width=9,echo=TRUE,warning=FALSE}
target_genes <- dge_tcell_csf[dge_tcell_csf$cluster %in% c("Expanded CD8 T cells","Non-expanded CD8 T cells"),]

# omiit Ribosomal genes (RP)
rp_genes     <- target_genes$gene[grep("^RP",target_genes$gene)]
target_genes <- target_genes[! target_genes$gene %in% rp_genes,]

# Most of the expression should be in the target cluster
target_genes <- target_genes[target_genes$pct.1 > 0.5,]

# Convert gene names to Entrez IDs
de <- as.character(mapIds(org.Hs.eg.db, target_genes$gene,'ENTREZID','SYMBOL'))
de <- de[!is.na(de)]

tcell_pathways <- enrichPathway(gene=de,pvalueCutoff=0.05, readable=T)
tcell_pathways_df <- as.data.frame(tcell_pathways)

barplot(tcell_pathways, showCategory=10)
```

#### Network of genes shared between pathways
```{r,fig.height=7, fig.width=15,echo=TRUE,warning=FALSE}
tcell_pathways_net <- pairwise_termsim(tcell_pathways,showCategory = 10)
emapplot(tcell_pathways_net, showCategory = 10)
```


# SAVE ENV - Temporary
```{r}
save(list = ls(),file = "objects/TEMP_ENV.RData")
```
