---
title: "Bonus Repertoire Exercises"
author: "BMS 270 Computational Immunology Workshop"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  markdown:
    wrap: 78
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")
```

## Purpose

These bonus exercises extend `3_repertoire_analysis.Rmd` with additional
immune-repertoire analyses that come up routinely in modern immunology papers
but are not covered in the core workshop. Each section is independent — feel
free to skip around.

Topics:

1.  **CDR3 physicochemical properties** (hydrophobicity, charge, polarity).
2.  **Public vs private clonotypes** across patients and compartments.
3.  **Compartment overlap with Morisita-Horn and Jaccard** indices.
4.  **Spectratype plots** — CDR3 length distribution per V gene.
5.  **CSF enrichment volcano** — clone-level log2(CSF/PB) vs abundance.
6.  **Isotype class-switching summary** for BCRs.
7.  **Expanded-vs-unexpanded gene-module scoring** (re-uses the Seurat
    object).

### Prior to running this code

```{r libs}
suppressPackageStartupMessages({
  library(alakazam); library(shazam)
  library(Seurat); library(dittoSeq)
  library(dplyr); library(tidyr); library(tibble); library(stringr)
  library(data.table)
  library(ggplot2); library(patchwork); library(ggrepel); library(scales)
  library(RColorBrewer); library(pheatmap)
})

source("resources/functions_workshop.R")
source("resources/functions_pub_networks.R")   # palettes + theme_pub()

load("objects/processed_objects.RData")

# Add patient + group columns
bcr_df$patient <- tstrsplit(bcr_df$sample, "_")[[1]]
tcr_df$patient <- tstrsplit(tcr_df$sample, "_")[[1]]
bcr_df$group   <- paste(bcr_df$patient, bcr_df$COMPARTMENT, sep = "_")
tcr_df$group   <- paste(tcr_df$patient, tcr_df$COMPARTMENT, sep = "_")

# Useful subsets
bcr_igh <- bcr_df[bcr_df$locus %in% "IGH", ]
tcr_trb <- tcr_df[tcr_df$locus %in% "TRB", ]

# Color schemes
status_pal     <- c(Healthy = "#009E73", Disease = "#CC79A7")
comp_pal       <- c(CSF = "#0072B2", PB = "#D55E00")
group_pal      <- c(`28_CSF` = "#08306B", `28_PB` = "#6BAED6",
                    `32_CSF` = "#7F2704", `32_PB` = "#FDAE6B")
```

### -----------------------------

# 1. CDR3 physicochemical properties

`alakazam::aminoAcidProperties()` computes per-sequence physicochemical
properties of a CDR3: length, hydrophobicity, charge, polarity, basicity,
acidity, aromaticity, isoelectric point (pI), and bulkiness. These properties
are often informative because antigen-experienced repertoires drift toward
hydrophobic, longer, more positively charged CDR3 loops (especially
CSF-resident expanded clones).

### a. Compute per-cell CDR3 properties

```{r cdr3_props}
# BCR heavy chain
bcr_props <- alakazam::aminoAcidProperties(
  bcr_igh, seq = "junction", trim = FALSE,
  property = c("length", "gravy", "polarity", "charge", "basic", "acidic", "aromatic")
)

# TCR beta chain
tcr_props <- alakazam::aminoAcidProperties(
  tcr_trb, seq = "junction", trim = FALSE,
  property = c("length", "gravy", "polarity", "charge", "basic", "acidic", "aromatic")
)
```

### b. Distribution of CDR3 hydrophobicity (GRAVY) and charge

```{r cdr3_grids, fig.height=4, fig.width=10}
# Compose a long-format table
to_long <- function(df, assay) {
  df %>%
    dplyr::select(unique_cell_id, COMPARTMENT, STATUS, SUBTYPE_ANNOTATION,
                  junction_aa_length, junction_aa_gravy, junction_aa_charge) %>%
    tidyr::pivot_longer(c(junction_aa_length, junction_aa_gravy,
                           junction_aa_charge),
                        names_to = "prop", values_to = "value") %>%
    dplyr::mutate(assay = assay,
                  prop  = dplyr::recode(prop,
                                        junction_aa_length = "CDR3 length (aa)",
                                        junction_aa_gravy  = "GRAVY (hydrophobicity)",
                                        junction_aa_charge = "Net charge"))
}

long_props <- dplyr::bind_rows(
  to_long(tcr_props, "TCR-β"),
  to_long(bcr_props, "BCR-IGH")
)

ggplot(long_props, aes(x = COMPARTMENT, y = value, fill = STATUS)) +
  geom_violin(scale = "width", color = "grey20", linewidth = 0.3,
              alpha = 0.85, position = position_dodge(0.7)) +
  geom_boxplot(width = 0.15, outlier.size = 0.4, color = "grey15",
               linewidth = 0.3, position = position_dodge(0.7)) +
  facet_grid(prop ~ assay, scales = "free_y") +
  scale_fill_manual(values = status_pal) +
  theme_pub() +
  labs(title = "CDR3 physicochemical properties by compartment and status",
       x = NULL, y = NULL, fill = "Status")
```

### c. Quick statistical test

```{r cdr3_test}
# Disease CSF vs PB charge — one example
disease_charge_csf <- tcr_props$junction_aa_charge[
  tcr_props$STATUS %in% "Disease" & tcr_props$COMPARTMENT %in% "CSF"]
disease_charge_pb  <- tcr_props$junction_aa_charge[
  tcr_props$STATUS %in% "Disease" & tcr_props$COMPARTMENT %in% "PB"]
wilcox.test(disease_charge_csf, disease_charge_pb)
```

### -----------------------------

# 2. Public vs private clonotypes

A *public* clone is shared across multiple patients (same CDR3-AA / V / J
identity), a *private* clone is unique to a single patient. In an autoimmune
disease setting, the public-clone fraction of expanded CSF cells is a
sensitive measure of antigen-driven convergent selection.

### a. Define clone identity

```{r public_priv_setup}
# Define an inter-patient clone key (CDR3-AA-beta + V-gene + J-gene)
tcr_pub <- tcr_trb %>%
  dplyr::filter(!is.na(junction_aa), junction_aa != "") %>%
  dplyr::mutate(
    clone_key = paste(junction_aa,
                       tstrsplit(v_call, "\\*")[[1]],
                       tstrsplit(j_call, "\\*")[[1]], sep = "|")
  ) %>%
  dplyr::distinct(unique_cell_id, clone_key, patient, COMPARTMENT, STATUS)

# How many patients does each clone_key appear in?
clone_breadth <- tcr_pub %>%
  dplyr::group_by(clone_key) %>%
  dplyr::summarize(n_patients = dplyr::n_distinct(patient),
                   n_cells    = dplyr::n_distinct(unique_cell_id),
                   .groups = "drop") %>%
  dplyr::mutate(category = ifelse(n_patients >= 2, "Public", "Private"))

table(clone_breadth$category)
```

### b. Plot public vs private cell fractions

```{r public_priv_plot, fig.height=4, fig.width=8}
cell_cat <- tcr_pub %>%
  dplyr::left_join(clone_breadth[, c("clone_key", "category")], by = "clone_key") %>%
  dplyr::count(STATUS, COMPARTMENT, category) %>%
  dplyr::group_by(STATUS, COMPARTMENT) %>%
  dplyr::mutate(frac = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(cell_cat, aes(x = COMPARTMENT, y = frac, fill = category)) +
  geom_col(width = 0.65, color = "grey20", linewidth = 0.3) +
  facet_wrap(~ STATUS) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c(Public = "#CC79A7", Private = "grey75")) +
  labs(title = "Public vs private TCR cells by compartment and status",
       x = NULL, y = "Fraction of cells", fill = "Clone type") +
  theme_pub()
```

### -----------------------------

# 3. Compartment overlap — Morisita-Horn and Jaccard

Two repertoires share clones to varying degrees. The Jaccard index measures
overlap by clone identity only (binary), while the Morisita-Horn index also
weights by abundance.

```{r overlap_funcs}
jaccard_index <- function(a, b) {
  a <- unique(a); b <- unique(b)
  length(intersect(a, b)) / length(union(a, b))
}

morisita_horn <- function(va, vb) {
  # va, vb: named numeric vectors (counts) keyed by clone id
  keys <- union(names(va), names(vb))
  va_full <- setNames(rep(0, length(keys)), keys); va_full[names(va)] <- va
  vb_full <- setNames(rep(0, length(keys)), keys); vb_full[names(vb)] <- vb
  na <- sum(va_full); nb <- sum(vb_full)
  if (na == 0 || nb == 0) return(NA_real_)
  num  <- 2 * sum(va_full * vb_full)
  den  <- (sum(va_full^2) / na^2 + sum(vb_full^2) / nb^2) * na * nb
  if (den == 0) return(NA_real_)
  num / den
}
```

### a. Per-(patient × compartment) overlap matrix

```{r overlap_matrix, fig.height=4, fig.width=10}
# Build a long count table
count_long <- tcr_trb %>%
  dplyr::distinct(unique_cell_id, clone_id, patient, COMPARTMENT) %>%
  dplyr::mutate(group = paste(patient, COMPARTMENT, sep = "_")) %>%
  dplyr::group_by(group, clone_id) %>%
  dplyr::summarize(n = dplyr::n_distinct(unique_cell_id), .groups = "drop")

groups <- unique(count_long$group)
mh_mat <- matrix(NA_real_, nrow = length(groups), ncol = length(groups),
                 dimnames = list(groups, groups))
jc_mat <- mh_mat

for (i in seq_along(groups)) {
  for (j in seq_along(groups)) {
    gi <- count_long[count_long$group == groups[i], ]
    gj <- count_long[count_long$group == groups[j], ]
    vi <- setNames(gi$n, gi$clone_id)
    vj <- setNames(gj$n, gj$clone_id)
    mh_mat[i, j] <- morisita_horn(vi, vj)
    jc_mat[i, j] <- jaccard_index(gi$clone_id, gj$clone_id)
  }
}

mh_df <- as.data.frame(as.table(mh_mat)); names(mh_df) <- c("g1","g2","value")
mh_df$metric <- "Morisita-Horn"
jc_df <- as.data.frame(as.table(jc_mat)); names(jc_df) <- c("g1","g2","value")
jc_df$metric <- "Jaccard"
ovl_df <- dplyr::bind_rows(mh_df, jc_df)

ggplot(ovl_df, aes(x = g1, y = g2, fill = value)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.3f", value)), size = 2.7, color = "grey15") +
  facet_wrap(~ metric) +
  scale_fill_gradientn(colors = c("#FFFFFF", "#C6DBEF", "#6BAED6", "#08519C"),
                       limits = c(0, max(ovl_df$value, na.rm = TRUE)),
                       name   = "Overlap") +
  coord_fixed() +
  labs(title = "TCR-β compartment overlap (Morisita-Horn & Jaccard)",
       x = NULL, y = NULL) +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())
```

### -----------------------------

# 4. Spectratype — CDR3 length distribution per V gene

Spectratype plots are a classic way to spot V-gene-restricted clonal
expansions. A polyclonal V gene shows a roughly normal CDR3-length
distribution; oligoclonal expansions show distorted spikes.

```{r spectratype, fig.height=4.5, fig.width=10}
spectra <- tcr_trb %>%
  dplyr::mutate(v_gene = tstrsplit(v_call, "\\*")[[1]],
                cdr3_len = nchar(as.character(junction_aa))) %>%
  dplyr::filter(!is.na(cdr3_len)) %>%
  dplyr::group_by(v_gene, cdr3_len, COMPARTMENT, STATUS) %>%
  dplyr::summarize(n = dplyr::n(), .groups = "drop")

# Pick the top 6 V genes overall for clarity
top_v <- spectra %>%
  dplyr::group_by(v_gene) %>%
  dplyr::summarize(total = sum(n), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(total)) %>%
  dplyr::slice_head(n = 6) %>%
  dplyr::pull(v_gene)

ggplot(spectra %>% dplyr::filter(v_gene %in% top_v),
       aes(x = cdr3_len, y = n, fill = COMPARTMENT)) +
  geom_col(position = "dodge", color = "grey20", linewidth = 0.2) +
  facet_grid(STATUS ~ v_gene, scales = "free_y") +
  scale_fill_manual(values = comp_pal) +
  labs(title = "TCR-β spectratype — top 6 V genes",
       x = "CDR3 length (aa)", y = "Cell count", fill = "Compartment") +
  theme_pub()
```

### -----------------------------

# 5. CSF-enrichment volcano (clone-level)

Each dot is a clonotype; the x-axis is log2(CSF / PB) clone fraction
(pseudocounts to avoid -Inf); the y-axis is total cell count. The top-right
quadrant is "expanded *and* CSF-enriched" — exactly the clones you would
prioritize for monoclonal antibody / TCR cloning.

```{r csf_volcano, fig.height=5, fig.width=8}
# Disease patient only — CSF/PB enrichment per clone
cn <- clone_counts_tcr
cn$log2_csf_pb   <- log2((cn$pct_CSF + 0.01) / (cn$pct_PB + 0.01))
cn$total_cells   <- cn$cell_count
cn$expanded     <- cn$cell_count >= 3 & cn$pct_CSF >= 0.1
cn$clone_id      <- rownames(cn)

ggplot(cn, aes(x = log2_csf_pb, y = total_cells, color = expanded)) +
  geom_hline(yintercept = 3, linetype = "dotted", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_color_manual(values = c(`TRUE` = "#D55E00", `FALSE` = "grey70"),
                     labels = c("Other", "Expanded & CSF-enriched"),
                     name = NULL) +
  scale_y_continuous(trans = "log1p", breaks = c(1, 3, 5, 10, 25, 50, 100)) +
  ggrepel::geom_text_repel(
    data = subset(cn, expanded & log2_csf_pb > 0),
    aes(label = clone_id),
    size = 2.6, max.overlaps = 25, min.segment.length = 0
  ) +
  labs(title = "CSF enrichment of TCR clones (Disease patient)",
       subtitle = "Top-right = expanded clones over-represented in CSF — mAb/TCR candidates",
       x = "log2(CSF fraction / PB fraction)",
       y = "Total cells per clone (log1p)") +
  theme_pub()
```

### -----------------------------

# 6. BCR isotype switching summary

Class-switching from IgM/IgD to IgG/IgA/IgE is a marker of T-help-driven
affinity maturation. The Disease CSF compartment is expected to be enriched in
class-switched BCRs relative to PB.

```{r isotype_summary, fig.height=4.5, fig.width=10}
iso_df <- bcr_igh %>%
  dplyr::distinct(unique_cell_id, c_call, COMPARTMENT, STATUS, patient,
                  SUBTYPE_ANNOTATION) %>%
  dplyr::filter(c_call != "", !is.na(c_call)) %>%
  dplyr::mutate(class = dplyr::case_when(
    c_call %in% c("IGHM", "IGHD") ~ "Naive (IgM/IgD)",
    grepl("IGHG", c_call) ~ "IgG",
    grepl("IGHA", c_call) ~ "IgA",
    c_call == "IGHE" ~ "IgE",
    TRUE ~ "Other"
  ))

iso_summary <- iso_df %>%
  dplyr::count(STATUS, COMPARTMENT, class) %>%
  dplyr::group_by(STATUS, COMPARTMENT) %>%
  dplyr::mutate(frac = n / sum(n)) %>%
  dplyr::ungroup()

p_iso_bar <- ggplot(iso_summary,
                    aes(x = COMPARTMENT, y = frac, fill = class)) +
  geom_col(width = 0.65, color = "grey20", linewidth = 0.3) +
  facet_wrap(~ STATUS) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("Naive (IgM/IgD)" = "#56B4E9",
                               "IgG" = "#E69F00",
                               "IgA" = "#009E73",
                               "IgE" = "#882255",
                               "Other" = "grey75")) +
  labs(title = "BCR isotype composition by compartment",
       x = NULL, y = "Fraction of B cells", fill = "Class") +
  theme_pub()

# Class-switched fraction = 1 - naive
p_csw <- iso_summary %>%
  dplyr::group_by(STATUS, COMPARTMENT) %>%
  dplyr::summarize(class_switched_frac = sum(frac[class != "Naive (IgM/IgD)"]),
                   .groups = "drop") %>%
  ggplot(aes(x = COMPARTMENT, y = class_switched_frac, fill = STATUS)) +
  geom_col(position = position_dodge(0.7), width = 0.6,
           color = "grey20", linewidth = 0.3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = status_pal) +
  labs(title = "Class-switched B cell fraction",
       x = NULL, y = "Class-switched (%)", fill = "Status") +
  theme_pub()

(p_iso_bar | p_csw) + plot_layout(widths = c(2, 1))
```

### -----------------------------

# 7. Gene-module scoring — expanded vs unexpanded clones

`Seurat::AddModuleScore()` lets you ask "do cells in expanded clones express a
coherent program?" The classic modules are cytotoxic effector, type-I
interferon, and tissue residency. Below we score a small effector module on
the T cell object and compare expanded vs non-expanded clones.

```{r module_score, fig.height=4, fig.width=8}
# Re-collapse expansion to Expanded vs Non-expanded
tcell_obj$expansion_two <- tcell_obj$expansion_category
tcell_obj$expansion_two[tcell_obj$expansion_two %in% c("High", "Moderate")] <- "Expanded"

# Effector module (Wherry / GTEx-curated cytotoxic effectors)
effector_genes <- list(c("GZMB","GZMA","GZMK","GZMH","PRF1","NKG7","GNLY","IFNG"))
exh_genes      <- list(c("PDCD1","HAVCR2","LAG3","TIGIT","TOX","CTLA4","ENTPD1"))
trm_genes      <- list(c("ITGAE","ITGA1","CXCR6","CD69","RGS1","ZNF683"))

tcell_obj <- Seurat::AddModuleScore(tcell_obj, features = effector_genes,
                                    name = "Effector_score",  ctrl = 50)
tcell_obj <- Seurat::AddModuleScore(tcell_obj, features = exh_genes,
                                    name = "Exhaustion_score", ctrl = 50)
tcell_obj <- Seurat::AddModuleScore(tcell_obj, features = trm_genes,
                                    name = "Trm_score",        ctrl = 50)

mod_df <- data.frame(
  cell        = colnames(tcell_obj),
  effector    = tcell_obj$Effector_score1,
  exhaustion  = tcell_obj$Exhaustion_score1,
  trm         = tcell_obj$Trm_score1,
  expansion   = tcell_obj$expansion_two,
  compartment = tcell_obj$compartment,
  status      = tcell_obj$status,
  subtype     = tcell_obj$sub.celltype.annot
) %>%
  dplyr::filter(expansion %in% c("Expanded", "Non-expanded"))

mod_long <- mod_df %>%
  tidyr::pivot_longer(c(effector, exhaustion, trm),
                       names_to = "module", values_to = "score") %>%
  dplyr::mutate(module = factor(module,
                                 levels = c("effector", "exhaustion", "trm"),
                                 labels = c("Effector", "Exhaustion",
                                            "Tissue residency (Trm)")))

ggplot(mod_long,
       aes(x = expansion, y = score, fill = expansion)) +
  geom_violin(scale = "width", color = "grey20", linewidth = 0.3, alpha = 0.85) +
  geom_boxplot(width = 0.15, outlier.size = 0.4,
               color = "grey15", linewidth = 0.3,
               fill = "white", alpha = 0.6) +
  facet_grid(module ~ status + compartment, scales = "free_y") +
  scale_fill_manual(values = c(Expanded = "#D55E00", `Non-expanded` = "grey75"),
                    guide = "none") +
  labs(title = "T cell module scores — expanded vs non-expanded",
       x = NULL, y = "Module score (AddModuleScore)") +
  theme_pub()
```

### -----------------------------

# Session info

```{r session_info}
sessionInfo()
```
