---
title: "Identify mAb/mTCR Candidates and Perform Network Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* Before starting set working directory to the CompImmunologyWorkshopUCSF folder
  * setwd("[PATH]/CompImmunologyWorkshopUCSF")


```{r,warning=FALSE}
# Set Working Directory before starting

# Immcantation
library(alakazam)
library(shazam)
library(tigger)
library(dowser)

library(Seurat)
library(dittoSeq)
library(reshape2)
library(kableExtra)
library(devtools)
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(gridExtra)    # grid.arrange()
library(dplyr)
library(impactSingleCellToolkit)

# GLIPH2 results directory
results_dir_gliph2 <- "4_FindCandidates/all_CSF_GLIPH2"

source("resources/functions_workshop.R")
```


# 1. Load Processed Objects
```{r}
load("objects/processed_objects.RData")
```

# -----------------------------

# Immune Repertoires


# -----------------------------



# 2. Dot plot of clonotypes

* Search for Disease Relevant Clonotypes (Thresholds based on prior analysis)
    * CSF_PB_RATIO_THRESH > 2
    * PB_count <= 1

```{r}
# Add column of CSF expanded/Other
CSF_PB_RATIO_THRESH = 2

# TCR
csf_high <- clone_counts_tcr[clone_counts_tcr$PB_count <= 1 & clone_counts_tcr$CSF_PB_ratio >= CSF_PB_RATIO_THRESH,]
csf_high_clonotypes <- row.names(csf_high)

exp_category <- row.names(clone_counts_tcr)
exp_category[exp_category %in% csf_high_clonotypes]   <- "CSF_high_expanded"
exp_category[! exp_category %in% "CSF_high_expanded"] <- "Other"
clone_counts_tcr$exp_category <- exp_category

# BCR
csf_high <- clone_counts_bcr[clone_counts_bcr$PB_count <= 1 & clone_counts_bcr$CSF_PB_ratio >= CSF_PB_RATIO_THRESH,]
csf_high_clonotypes <- row.names(csf_high)

exp_category <- row.names(clone_counts_bcr)
exp_category[exp_category %in% csf_high_clonotypes]   <- "CSF_high_expanded"
exp_category[! exp_category %in% "CSF_high_expanded"] <- "Other"
clone_counts_bcr$exp_category <- exp_category
```

* Most B cell clonotypes are 1 cell either in the CSF or PB
```{r,fig.height=5, fig.width=15,echo=FALSE,warning=FALSE}
# TCR 
p1 <- ggplot(clone_counts_tcr, aes(x=pct_PB, y=pct_CSF,color=exp_category)) + geom_point(aes(size=cell_count)) + scale_color_manual(breaks = c("CSF_high_expanded", "Other"),values=c("red", "black")) + ggtitle("T cells")

p2 <- ggplot(clone_counts_bcr, aes(x=pct_PB, y=pct_CSF,color=exp_category)) + geom_point(aes(size=cell_count)) + scale_color_manual(breaks = c("CSF_high_expanded", "Other"),values=c("red", "black")) + ggtitle("B cells")

#p1 <- p1 + scale_x_continuous(trans = 'log2') + scale_y_continuous(trans = 'log2')
grid.arrange(p1,p2, nrow = 1)
```

# -----------------------------

# GLIPH2 Network Analysis

# 1. Generate input files
* Running GLIPH2 is optional, results files provided
* generateGLIPH2inputFiles will create all the input files you need for GLIPH2, but running GLIPH2 still requires a parameter plain text file (made manually)

```{r}
tcr_df_csf         <- tcr_df[tcr_df$COMPARTMENT %in% "CSF",]
tcr_df_csf$patient <- tstrsplit(tcr_df_csf$sample,"_")[[1]]

generateGLIPH2inputFiles(tcr_df_csf,results.dir = results_dir_gliph2)
```


# 2. Run GLIPH2 in the terminal (OPTIONAL)

### a. Format input parameter file
* parameter files have been provided as part of the workshop
* Example: parameter file will include the following lines:
```{bash}
out_prefix=AllCSF
cdr3_file=TCR_input_table.GLIPH2.tsv
refer_file=ref.txt
v_usage_freq_file=ref_V.txt
cdr3_length_freq_file=ref_L.txt
local_min_pvalue=0.001
p_depth=1000
global_convergence_cutoff=1
simulation_depth=1000
kmer_min_depth=3
local_min_OVE=10
algorithm=GLIPH2
all_aa_interchangeable=1
```

* To inlcude HLA genotypes, create a TSV with all class 1 and 2 alleles (1 row per patient)
* Example: adding HLA geneotypes in parameter file:
```{bash}
hla_file=HLA_table_formatted_GLIPH.class1and2.tab
```

### b. Run GLIPH2 in terminal (OPTIONAL)
* Results provided used input patient HLA genotypes
```{bash}
cd 4_FindCandidates/all_CSF_GLIPH2
irtools.centos -c param_file_gliph2.cfg
```


# 3. Load GLIPH2 results for all CSF clones
```{r}
gliph_results_fh <- "4_FindCandidates/all_CSF_GLIPH2/AllCSF_cluster.csv"
gliph_resuls_csf <- read.csv(gliph_results_fh,header = T,stringsAsFactors = F)
gliph_resuls_csf$CRD3ab_AA <- paste(gliph_resuls_csf$TcRa,gliph_resuls_csf$TcRb,sep = ":")
all_groups       <- unique(gliph_resuls_csf$index)
```

# 2. Add CSF clonotype frequencies and cell counts to TCR results
```{r}
tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_results,clone_counts = clone_counts_tcr,target.col = "pct_CSF",new.col.name = "clonotype_CSF_freq")
tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_results_gliph,clone_counts = clone_counts_tcr,target.col = "cell_count",new.col.name = "cell_count")
tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_results_gliph,clone_counts = clone_counts_tcr,target.col = "CSF_count",new.col.name = "CSF_count")
```

# 3. Add GLIPH results into mAb table

### a. GLIPH status 
```{r}
# Add Unified CDR3 Alpha/Beta string to TCR table
tcr_cdr3_col <- tcr_results_gliph$CELL_BARCODES

for (cell in unique(tcr_results_gliph$CELL_BARCODES)) {
  cell_df <- tcr_results_gliph[tcr_results_gliph$CELL_BARCODES %in% cell,]
  alpha   <- cell_df[cell_df$locus %in% "TRA",]
  beta    <- cell_df[cell_df$locus %in% "TRB",]
  ab      <- paste(c(alpha$junction_aa,beta$junction_aa),collapse = ":")
  tcr_cdr3_col[tcr_cdr3_col %in% cell] <- ab
}
tcr_results_gliph$CDR3ab_AA <- tcr_cdr3_col

# Gliph results status 
gliph_status_col <- tcr_results_gliph$CDR3ab_AA
gliph_status_col[gliph_status_col %in% unique(gliph_resuls_csf$CRD3ab_AA)] <- "1"
gliph_status_col[! gliph_status_col %in% "1"] <- "0"
tcr_results_gliph$gliph_status <- gliph_status_col
```

### b. Add index groups
```{r}
target_cells <- tcr_results_gliph[tcr_results_gliph$CDR3ab_AA %in% gliph_resuls_csf$CRD3ab_AA,]
target_cells <- unique(target_cells$CELL_BARCODES)

index_col        <- tcr_results_gliph$CDR3ab_AA
names(index_col) <- tcr_results_gliph$CELL_BARCODES
index_col[! names(index_col) %in% target_cells] <- ""

target_cdr3ab <- unique(as.character(index_col[! index_col %in% ""]))


for (ab in target_cdr3ab) {
  gliph_sub <- gliph_resuls_csf[gliph_resuls_csf$CRD3ab_AA %in% ab,]
  if(nrow(gliph_sub) > 0){
    index_vec <- unique(gliph_sub$index)
    if(length(index_vec) > 1){
      index_fmt <- paste(sort(index_vec),collapse = ":")
      index_col[index_col %in% ab] <- index_fmt
    } else{
      index_col[index_col %in% ab] <- index_vec
    }
    
  }
}

tcr_results_gliph$gliph_group <- index_col
```


# 4. Create GLIPH Network
```{r}
clonotype_counts_csf     <- clone_counts_tcr[clone_counts_tcr$CSF_count > 1,]
clonotype_counts_csf_exp <- clonotype_counts_csf[clonotype_counts_csf$cell_count > 1,]
THRESH <- 0.75
high_csf_clones     <- clonotype_counts_csf_exp[clonotype_counts_csf_exp$CSF_PB_ratio > THRESH,]
moderate_csf_clones <- clonotype_counts_csf_exp[clonotype_counts_csf_exp$CSF_PB_ratio <= THRESH,]


target_clones <- row.names(high_csf_clones) # Expanded CSF clones

# Omit cells with no Alpha CDR3
tcr_results_gliph$junction_aa
data_tcr_na     <- tcr_results_gliph[is.na(tcr_results_gliph$junction_aa),]
incomplete_tcrs <- unique(data_tcr_na$CELL_BARCODES)
data_tcr_clean  <- tcr_results_gliph[! tcr_results_gliph$CELL_BARCODES %in% incomplete_tcrs,]

# Get target CSF high Clonotypes
data_tcr_csf     <- data_tcr_clean[data_tcr_clean$COMPARTMENT %in% c("CSF"),]
data_tcr_csf_exp <- data_tcr_csf[data_tcr_csf$clone_id %in% row.names(clonotype_counts_csf_exp),]
data_tcr_csf_exp <- data_tcr_csf_exp[data_tcr_csf_exp$gliph_status %in% "1",]

data_tcr_target <- data_tcr_csf_exp[data_tcr_csf_exp$clone_id %in% target_clones,]

# Isoloate all cells with interacting CDR3s (all GLIPH 2 groups)
target_groups <- unique(data_tcr_target$gliph_group)

target_groups_all <- c()
for (grp in target_groups) {
  if(grepl(":",grp)){
    # string contains more than one TCR group
    # Add in full grp interaction string
    target_groups_all <- c(target_groups_all,grp)
    
    # Add in all individual groups within interaction string
    grp_list <- unlist(strsplit(grp, ":"))
    target_groups_all <- c(target_groups_all,grp_list)
  } else{
    target_groups_all <- c(target_groups_all,grp)
  }
}
target_groups_all <- unique(target_groups_all)

interac_tcr <- data_tcr_csf[data_tcr_csf$gliph_group %in% target_groups_all,]
interac_tcr <- unique(interac_tcr$CELL_BARCODES)

# Get all CSF high and cells with interacting TCRs
all_target_cells <- c(unique(data_tcr_target$CELL_BARCODES),interac_tcr)
data_tcr_target_all <- data_tcr_clean[data_tcr_clean$CELL_BARCODES %in% all_target_cells,]


# Omit nodes with no interactions (gliph  interactions = 0)
data_tcr_target_all <- data_tcr_target_all[data_tcr_target_all$gliph_status %in% "1",]


# Create GML file - Clonotype level network
createGLIPH2clonotypeNetworkGML(
  data_tcr   =  data_tcr_target_all,
  result_dir = "4_FindCandidates/CSF_High_ClonotypeNetwork_GLIPH2",
  output_name     = "gliph_cleaned_clusters_CSF_high_GLIPH2",
  count_cutoff    =  as.numeric(2),  # UMI Count cut-off
  distance_cutoff =  as.numeric(0),
  method = "hamming",
  cores = 4)
```


# 5. Create TCR clonotype connection network
```{r}
tcr_results_gliph_sub <- tcr_results_gliph[tcr_results_gliph$cell_count > 1,]
createTCRnetworkGML(
  data_tcr   =  tcr_results_gliph_sub,
  result_dir = "4_FindCandidates/TCR_ClonotypeNetwork",
  output_name     = "cleaned_clusters",
  count_cutoff    =  as.numeric(2),  # UMI Count cut-off
  distance_cutoff =  as.numeric(0),
  method = "hamming",
  cores = 4)


```


# 6. Format Results for a Query in VDJdb (CSF only)

format CSF TCR results for upload to VDJ db

### a. add clonotype frequencies to TCR CSF results
```{r}
# Transfer clonotype Freq back to full results
tcr_results_csf <- tcr_results_gliph[tcr_results_gliph$COMPARTMENT %in% "CSF",]
tcr_results_csf <- tcr_results_csf[tcr_results_csf$STATUS %in% "Disease",]
clonotype_freq    <- tcr_results_csf$clone_id
clonotype_count   <- tcr_results_csf$clone_id

for(clone in row.names(clone_counts_tcr)){
  clone_row  <- clone_counts_tcr[row.names(clone_counts_tcr) %in% clone,]
  cell_count <- clone_row$CSF_count
  clone_freq <- clone_row$pct_CSF / 100
  
  clonotype_count[clonotype_count %in% clone] <- cell_count
  clonotype_freq[clonotype_freq %in% clone]   <- clone_freq
}

tcr_results_csf$CLONE_COUNT <- clonotype_count
tcr_results_csf$CLONE_FREQ  <- clonotype_freq
```

### b. format table to meet VDJdb format
```{r}
cols_to_keep  <- c("CLONE_COUNT","CLONE_FREQ","junction","junction_aa","v_call","d_call","j_call")
fmt_col_names <- c("count","frequency","CDR3nt","CDR3aa","V","D","J")

fmt_tcr_results <- tcr_results_csf[,cols_to_keep]
names(fmt_tcr_results) <- fmt_col_names

fmt_tcr_results$D      <- "."
fmt_tcr_results$Vend   <- ""
fmt_tcr_results$Dstart <- ""
fmt_tcr_results$Dend   <- ""
fmt_tcr_results$Jstart <- ""

```

# 7. Output table and upload to VDJdb

VDJdb <https://vdjdb.cdr3.net/>
1. You will need to create an account or login temporarily.
2. Once logged in, click on the 'Annotation' tab
3. Click on the 'Upload' button
4. Click 'Add' and select the file we are generating bellow 
```{r}
write.table(fmt_tcr_results,file = "4_FindCandidates/table_for_VDJdb_CSF.txt",quote = F,row.names = F,sep = "\t")
```


# 8. Assembling monoclonals

### a. BCR - assemble FR and CDRs for clonotypes of interest
```{r}
# B cells
target_bcr <- clone_counts_bcr[clone_counts_bcr$exp_category %in% "CSF_high_expanded",]
target_bcr <- row.names(target_bcr)[1]

# subset all cells in target clone (Heavy and Light chains)
bcr_results$fmt_cell_id <- tstrsplit(bcr_results$sequence_id,"_contig")[[1]]
target_bcr_df <- bcr_results[bcr_results$clone_id %in% target_bcr,] # heavy chains only
target_bcr_df <- bcr_results[bcr_results$fmt_cell_id %in% target_bcr_df$fmt_cell_id,]

# Subset columns
cols_to_keep <- c("fmt_cell_id","locus","v_call","d_call","j_call","clone_id",
                  "fwr1","cdr1","fwr2","cdr2","fwr3","cdr3","fwr4")
target_bcr_df <- target_bcr_df[,cols_to_keep]

# concatenate sequences and remove gaps
target_bcr_df$monoclonal_seq <- paste(target_bcr_df$fwr1,target_bcr_df$cdr1,
                                      target_bcr_df$fwr2,target_bcr_df$cdr2,
                                      target_bcr_df$fwr3,target_bcr_df$cdr3,target_bcr_df$fwr4,sep = "")
target_bcr_df$monoclonal_seq <- str_replace_all(target_bcr_df$monoclonal_seq,"\\.","")

target_bcr_df_sub <- target_bcr_df[,c("fmt_cell_id","locus","v_call","d_call","j_call","clone_id","monoclonal_seq")]
```

### b. TCR - assemble FR and CDRs for clonotypes of interest
```{r}
# T cells
target_tcr <- clone_counts_tcr[clone_counts_tcr$exp_category %in% "CSF_high_expanded",]
target_tcr <- target_tcr[order(target_tcr$CSF_PB_ratio,decreasing = T),]
target_tcr <- row.names(target_tcr)[1]

# subset all cells in target clone (Heavy and Light chains)
tcr_results$fmt_cell_id <- tstrsplit(tcr_results$sequence_id,"_contig")[[1]]
target_tcr_df <- tcr_results[tcr_results$clone_id %in% target_tcr,] # Alpha and Beta

# Subset columns
cols_to_keep <- c("fmt_cell_id","locus","v_call","d_call","j_call","clone_id",
                  "fwr1","cdr1","fwr2","cdr2","fwr3","cdr3","fwr4")
target_tcr_df <- target_tcr_df[,cols_to_keep]

# concatenate sequences and remove gaps
target_tcr_df$monoclonal_seq <- paste(target_tcr_df$fwr1,target_tcr_df$cdr1,
                                      target_tcr_df$fwr2,target_tcr_df$cdr2,
                                      target_tcr_df$fwr3,target_tcr_df$cdr3,target_tcr_df$fwr4,sep = "")
target_tcr_df$monoclonal_seq <- str_replace_all(target_tcr_df$monoclonal_seq,"\\.","")

target_tcr_df_sub <- target_tcr_df[,c("fmt_cell_id","locus","v_call","d_call","j_call","clone_id","monoclonal_seq")]
```

### c. Other tools

### --stichr (TCRs only)
* https://github.com/JamieHeather/stitchr
```{bash}
python3 stitchr.py -v TRBV7-6 -j TRBJ1-4 -cdr3 CASSSGQGLGEKLFF
```



