---
title: "Publication-Quality VDJ Repertoire Networks"
author: "BMS 270 Computational Immunology Workshop"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  markdown:
    wrap: 78
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  dev        = "cairo_pdf"
)
```

## Overview

This script takes the BCR / TCR repertoire data already produced by
**1_filter_rnaseq_and_vdj_data.Rmd** through **4_find_candidates.Rmd** and
re-renders the network figures at publication quality. The goal is figures
that could go straight into a high-impact paper without further re-styling.

What this script adds compared to `4_find_candidates.Rmd`:

1.  **ggraph + tidygraph backend.** Networks are drawn through ggplot2's
    grammar, so layout, color, size, theme and legend all behave like any
    other ggplot.
2.  **Colorblind-safe palettes** (Okabe-Ito) used consistently across panels.
3.  **Multi-panel composition with patchwork.** Each Figure X is a single
    object you can write to a single PDF.
4.  **Companion visualizations** that contextualize the networks:
    - V-J gene-pairing chord diagrams (`circlize`)
    - Clone-size rank distributions (Zipf-style log-log plots)
    - Hill diversity profiles (q = 0 → 4) with bootstrap 95% CIs
5.  **PDF (vector) output at journal column widths** (85 / 115 / 180 mm) using
    the `cairo_pdf` device, ready for Adobe Illustrator / Inkscape touch-up.

### Prior to running this code

- Set the working directory to the **CompImmunologyWorkshopUCSF** folder:
  `setwd("[PATH]/CompImmunologyWorkshopUCSF")`
- This script depends on the same processed objects as `4_find_candidates.Rmd`
  (i.e. `objects/processed_objects.RData`). It is otherwise self-contained.

### Additional packages required

```{r install_check, eval=FALSE}
# Run once to make sure the publication-figure stack is installed
pkgs <- c("tidygraph", "ggraph", "graphlayouts", "patchwork", "circlize",
          "ggrepel", "ggnewscale", "scales", "ggsci", "tibble", "tidyr")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, repos = "https://cloud.r-project.org")
```

```{r libs, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  # Core data
  library(dplyr); library(tidyr); library(tibble); library(stringr)
  library(data.table)
  # Network / graph stack
  library(igraph); library(tidygraph); library(ggraph)
  # Plotting
  library(ggplot2); library(scales); library(patchwork)
  library(RColorBrewer); library(ggrepel); library(ggnewscale)
  # Existing workshop helpers
  library(impactSingleCellToolkit)
})

source("resources/functions_workshop.R")
source("resources/functions_pub_networks.R")

# Reproducibility
seed.number <- 1234
set.seed(seed.number)

# Where to write figures
fig_dir <- "figures"
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

# 1. Load processed objects

```{r load_data}
load("objects/processed_objects.RData")

# Add patient / status columns (mirrors 4_find_candidates.Rmd)
bcr_df$patient <- tstrsplit(bcr_df$sample, "_")[[1]]
tcr_df$patient <- tstrsplit(tcr_df$sample, "_")[[1]]

# Friendly labels
patient_label <- c(`28` = "Disease (P28)", `32` = "Healthy (P32)")
bcr_df$patient_label <- patient_label[bcr_df$patient]
tcr_df$patient_label <- patient_label[tcr_df$patient]
```

### -----------------------------

# 2. BCR repertoire networks (Figure 1)

We build one tidygraph per patient and lay them out side-by-side. Each node is
a B cell; each cluster of connected nodes is a clonotype. Node size is
log-scaled to the clone size; fill encodes compartment (CSF vs PB).

The layout is **category-constrained**: each clone is classified as CSF-pure,
compartment-shared, or PB-pure based on the composition of its member cells,
and clones are packed into three side-by-side bands accordingly. Within each
band clones are arranged in a non-overlapping grid; within each clone, cells
are laid out on a tight circle whose radius scales with √(clone size). The
result is a network where the spatial story is readable at a glance: "look
at the left band → those are CSF-restricted clones".

### a. Build nodes / edges / graphs

```{r bcr_graphs}
# Use the workshop helpers (same as 4_find_candidates.Rmd) so the underlying
# node/edge structure is identical -- we just re-render it nicely.
bcr_node_dis <- createRepertoireNodeDataframe(
  bcr_df, assay = "bcr", select.patients = "28", expansion.thresh = 2
)
# id format from the helper is "{barcode}-{patient}_{compartment}", with
# compartment encoded as PBMC. Normalize to PB to match the rest of the data.
bcr_node_dis$compartment <- normalize_compartment(tstrsplit(bcr_node_dis$id, "_")[[2]])
bcr_node_dis$patient     <- "28"
bcr_edges_dis <- createRepertoireEdgeDataframe(bcr_node_dis)
g_bcr_dis     <- build_repertoire_tidygraph(bcr_node_dis, bcr_edges_dis)

bcr_node_hc <- createRepertoireNodeDataframe(
  bcr_df, assay = "bcr", select.patients = "32", expansion.thresh = 2
)
bcr_node_hc$compartment <- normalize_compartment(tstrsplit(bcr_node_hc$id, "_")[[2]])
bcr_node_hc$patient     <- "32"
bcr_edges_hc <- createRepertoireEdgeDataframe(bcr_node_hc)
g_bcr_hc     <- build_repertoire_tidygraph(bcr_node_hc, bcr_edges_hc)
```

### b. Render each panel

```{r bcr_plots, fig.height=6, fig.width=13}
p_bcr_dis <- plot_repertoire_network(
  g_bcr_dis,
  color_by    = "compartment",
  layout      = "grouped",
  layout_args = list(node_radius = 0.030, gap_factor = 1.25, aspect = 1.4),
  edge_alpha  = 0.25,
  edge_width  = 0.30,
  size_range  = c(1.4, 5.5),
  label_top_n = 5,
  label_col   = "clone_id",
  band_labels = TRUE,
  title       = "B cell repertoire — Disease (P28)",
  subtitle    = "Left band: CSF-restricted   Middle: compartment-shared   Right: PB-restricted",
  seed        = seed.number
)

p_bcr_hc <- plot_repertoire_network(
  g_bcr_hc,
  color_by    = "compartment",
  layout      = "grouped",
  layout_args = list(node_radius = 0.030, gap_factor = 1.25, aspect = 1.4),
  edge_alpha  = 0.25,
  edge_width  = 0.30,
  size_range  = c(1.4, 5.5),
  label_top_n = 5,
  label_col   = "clone_id",
  band_labels = TRUE,
  title       = "B cell repertoire — Healthy (P32)",
  subtitle    = "Left band: CSF-restricted   Middle: compartment-shared   Right: PB-restricted",
  seed        = seed.number
)

fig_bcr <- compose_panels(list(p_bcr_dis, p_bcr_hc), ncol = 2,
                          tag_levels = "a")
fig_bcr
```

### c. Save vector PDF + 300-DPI PNG

```{r bcr_pdf, eval=FALSE}
save_pub_figure(fig_bcr,
                file_basename = file.path(fig_dir, "Fig1_BCR_networks"),
                width_class   = "double",
                height_mm     = 130,
                also_png      = TRUE)
```

### -----------------------------

# 3. TCR (expanded) repertoire networks (Figure 2)

T cell repertoires are much larger than BCR repertoires, so we restrict to
expanded clones (cell count \> 1) and split by CD4 vs CD8 in the next panel.

### a. Build graphs

```{r tcr_graphs}
tcr_df_exp <- tcr_df[! tcr_df$EXPANSION_CATEGORY %in% "Non-expanded", ]

# Disease
tcr_node_dis <- createRepertoireNodeDataframe(
  tcr_df_exp, assay = "tcr", select.patients = "28", expansion.thresh = 2
)
tcr_node_dis$compartment <- normalize_compartment(tstrsplit(tcr_node_dis$id, "_")[[2]])
tcr_node_dis$patient     <- "28"
tcr_edges_dis <- createRepertoireEdgeDataframe(tcr_node_dis)
g_tcr_dis     <- build_repertoire_tidygraph(tcr_node_dis, tcr_edges_dis)

# Healthy
tcr_node_hc <- createRepertoireNodeDataframe(
  tcr_df_exp, assay = "tcr", select.patients = "32", expansion.thresh = 2
)
tcr_node_hc$compartment <- normalize_compartment(tstrsplit(tcr_node_hc$id, "_")[[2]])
tcr_node_hc$patient     <- "32"
tcr_edges_hc <- createRepertoireEdgeDataframe(tcr_node_hc)
g_tcr_hc     <- build_repertoire_tidygraph(tcr_node_hc, tcr_edges_hc)
```

### b. Render

```{r tcr_plots, fig.height=6, fig.width=13}
p_tcr_dis <- plot_repertoire_network(
  g_tcr_dis,
  color_by    = "compartment",
  layout      = "grouped",
  layout_args = list(node_radius = 0.020, gap_factor = 1.20, aspect = 1.4),
  edge_alpha  = 0.18,
  edge_width  = 0.22,
  size_range  = c(1.0, 5.0),
  label_top_n = 4,
  band_labels = TRUE,
  title       = "T cell repertoire — Disease (P28)",
  subtitle    = "Expanded clones only (cell count > 1)",
  seed        = seed.number
)

p_tcr_hc <- plot_repertoire_network(
  g_tcr_hc,
  color_by    = "compartment",
  layout      = "grouped",
  layout_args = list(node_radius = 0.020, gap_factor = 1.20, aspect = 1.4),
  edge_alpha  = 0.18,
  edge_width  = 0.22,
  size_range  = c(1.0, 5.0),
  label_top_n = 4,
  band_labels = TRUE,
  title       = "T cell repertoire — Healthy (P32)",
  subtitle    = "Expanded clones only (cell count > 1)",
  seed        = seed.number
)

fig_tcr <- compose_panels(list(p_tcr_dis, p_tcr_hc), ncol = 2, tag_levels = "a")
fig_tcr
```

```{r tcr_pdf, eval=FALSE}
save_pub_figure(fig_tcr,
                file_basename = file.path(fig_dir, "Fig2_TCR_networks"),
                width_class   = "double",
                height_mm     = 130,
                also_png      = TRUE)
```

### -----------------------------

# 4. CD8 T cell networks colored by CDR3 length (Figure 3)

A small but high-information addition: color expanded CD8 clones by mean CDR3
length, which often tracks antigen-experience and tissue residency.

### a. Build graph with CDR3 metadata

```{r cd8_graph}
tcr_df_cd8 <- tcr_df_exp[tcr_df_exp$SUBTYPE_ANNOTATION %in% "CD8 T cells", ]
tcr_df_cd8 <- tcr_df_cd8[tcr_df_cd8$locus %in% "TRB", ]   # beta only for CDR3 length

# Mean beta CDR3 length per clone (one beta per cell after this filter)
tcr_df_cd8$cdr3_len <- nchar(as.character(tcr_df_cd8$junction_aa))
clone_cdr3_len <- tapply(tcr_df_cd8$cdr3_len, tcr_df_cd8$clone_id,
                         function(x) mean(x, na.rm = TRUE))

# Per-cell patient lookup so we don't depend on id-parsing
cell_patient <- setNames(tcr_df_cd8$patient, tcr_df_cd8$unique_cell_id)
cell_patient <- cell_patient[!duplicated(names(cell_patient))]

cd8_node <- createRepertoireNodeDataframe(
  tcr_df_cd8, assay = "tcr", select.patients = c("28", "32"),
  expansion.thresh = 2
)
cd8_node$compartment   <- normalize_compartment(tstrsplit(cd8_node$id, "_")[[2]])
cd8_node$patient       <- unname(cell_patient[cd8_node$id])
cd8_node$mean_cdr3_len <- as.numeric(clone_cdr3_len[as.character(cd8_node$clone_id)])

cd8_edges <- createRepertoireEdgeDataframe(cd8_node)
g_cd8     <- build_repertoire_tidygraph(cd8_node, cd8_edges)
```


### -----------------------------

# 5. GLIPH motif network (Figure 4)

GLIPH2 groups TCRs by shared CDR3 motifs that putatively bind the same
antigen. Here we re-render the GLIPH network with proper labels on the motif
hubs, colored by patient, sized by clone size.

### a. Re-run the same filtering as `4_find_candidates.Rmd`

```{r gliph_filter}
gliph_results_fh <- "4_FindCandidates/all_CSF_GLIPH2/AllCSF_cluster.csv"
gliph_resuls_csf <- read.csv(gliph_results_fh, header = TRUE, stringsAsFactors = FALSE)
gliph_resuls_csf$CRD3ab_AA <- paste(gliph_resuls_csf$TcRa, gliph_resuls_csf$TcRb, sep = ":")

tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_df,
                                                   clone_counts = clone_counts_tcr,
                                                   target.col = "pct_CSF",
                                                   new.col.name = "clonotype_CSF_freq")
tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_results_gliph,
                                                   clone_counts = clone_counts_tcr,
                                                   target.col = "cell_count",
                                                   new.col.name = "cell_count")
tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_results_gliph,
                                                   clone_counts = clone_counts_tcr,
                                                   target.col = "CSF_count",
                                                   new.col.name = "CSF_count")

# Build CDR3ab_AA per cell
tcr_cdr3_col <- tcr_results_gliph$unique_cell_id
for (cell in unique(tcr_results_gliph$unique_cell_id)) {
  cell_df <- tcr_results_gliph[tcr_results_gliph$unique_cell_id %in% cell, ]
  alpha   <- cell_df[cell_df$locus %in% "TRA", ]
  beta    <- cell_df[cell_df$locus %in% "TRB", ]
  ab      <- paste(c(alpha$junction_aa, beta$junction_aa), collapse = ":")
  tcr_cdr3_col[tcr_cdr3_col %in% cell] <- ab
}
tcr_results_gliph$CDR3ab_AA <- tcr_cdr3_col

# GLIPH membership flag
gliph_status_col <- tcr_results_gliph$CDR3ab_AA
gliph_status_col[gliph_status_col %in% unique(gliph_resuls_csf$CRD3ab_AA)] <- "1"
gliph_status_col[! gliph_status_col %in% "1"] <- "0"
tcr_results_gliph$gliph_status <- gliph_status_col

# Index group string per cell
target_cells <- tcr_results_gliph[tcr_results_gliph$CDR3ab_AA %in% gliph_resuls_csf$CRD3ab_AA, ]
target_cells <- unique(target_cells$unique_cell_id)

index_col        <- tcr_results_gliph$CDR3ab_AA
names(index_col) <- tcr_results_gliph$unique_cell_id
index_col[! names(index_col) %in% target_cells] <- ""
target_cdr3ab <- unique(as.character(index_col[! index_col %in% ""]))
for (ab in target_cdr3ab) {
  gliph_sub <- gliph_resuls_csf[gliph_resuls_csf$CRD3ab_AA %in% ab, ]
  if (nrow(gliph_sub) > 0) {
    index_vec <- unique(gliph_sub$index)
    if (length(index_vec) > 1) {
      index_col[index_col %in% ab] <- paste(sort(index_vec), collapse = ":")
    } else {
      index_col[index_col %in% ab] <- index_vec
    }
  }
}
tcr_results_gliph$gliph_group <- index_col

# Subset to expanded, GLIPH-positive CSF clones (mirrors 4_find_candidates.Rmd)
THRESH <- 0.75
clonotype_counts_csf     <- clone_counts_tcr[clone_counts_tcr$CSF_count > 1, ]
clonotype_counts_csf_exp <- clonotype_counts_csf[clonotype_counts_csf$cell_count > 1, ]
high_csf_clones <- clonotype_counts_csf_exp[clonotype_counts_csf_exp$CSF_PB_ratio > THRESH, ]
target_clones   <- row.names(high_csf_clones)

data_tcr_na     <- tcr_results_gliph[is.na(tcr_results_gliph$junction_aa), ]
incomplete_tcrs <- unique(data_tcr_na$unique_cell_id)
data_tcr_clean  <- tcr_results_gliph[! tcr_results_gliph$unique_cell_id %in% incomplete_tcrs, ]

data_tcr_csf     <- data_tcr_clean[data_tcr_clean$COMPARTMENT %in% "CSF", ]
data_tcr_csf_exp <- data_tcr_csf[data_tcr_csf$clone_id %in% row.names(clonotype_counts_csf_exp), ]
data_tcr_csf_exp <- data_tcr_csf_exp[data_tcr_csf_exp$gliph_status %in% "1", ]
data_tcr_target  <- data_tcr_csf_exp[data_tcr_csf_exp$clone_id %in% target_clones, ]

target_groups <- unique(data_tcr_target$gliph_group)
target_groups_all <- unique(unlist(strsplit(target_groups, ":")))
target_groups_all <- c(target_groups_all, target_groups)
target_groups_all <- unique(target_groups_all)

interac_tcr <- unique(data_tcr_csf[data_tcr_csf$gliph_group %in% target_groups_all,
                                    "unique_cell_id"])

all_target_cells     <- c(unique(data_tcr_target$unique_cell_id), interac_tcr)
data_tcr_target_all  <- data_tcr_clean[data_tcr_clean$unique_cell_id %in% all_target_cells, ]
data_tcr_target_all  <- data_tcr_target_all[data_tcr_target_all$gliph_status %in% "1", ]
```

### b. Build GLIPH-grouped node / edge / graph

```{r gliph_graph}
gliph_node <- createRepertoireNodeDataframe(
  data_tcr_target_all, assay = "tcr",
  clone.col       = "gliph_group",
  prefix.clone.id = "gliph",
  expansion.thresh = 2
)
row.names(gliph_node) <- gliph_node$id

# Add patient and clone-id metadata to nodes (one row per cell)
extra_df <- data_tcr_target_all[data_tcr_target_all$locus %in% "TRB", ]
extra_df <- extra_df[!duplicated(extra_df$unique_cell_id), ]
row.names(extra_df) <- extra_df$unique_cell_id
extra_df <- extra_df[, c("clone_id", "patient", "COMPARTMENT")]
names(extra_df)[1] <- "clone_id_vdj"

gliph_node <- merge(gliph_node, extra_df, by = 0, all.x = TRUE)
gliph_node[, 1] <- NULL
row.names(gliph_node) <- gliph_node$id

# Compact motif label: one label per GLIPH group, on the largest node
net_label_col <- gliph_node$clone_id
for (grp in unique(net_label_col)) {
  idx <- which(net_label_col %in% grp)
  # keep the label on the first row (one motif label per cluster is enough)
  if (length(idx) > 1) net_label_col[idx[-1]] <- ""
}
gliph_node$net_node_label <- net_label_col

gliph_edges <- createRepertoireEdgeDataframe(gliph_node)
g_gliph     <- build_repertoire_tidygraph(gliph_node, gliph_edges)
```

### c. Render — labels on motif hubs, grouped layout by patient

For the GLIPH network the meaningful spatial axis is *patient identity*
rather than compartment, since GLIPH groups are by definition CSF-only here.
We use `compartment_col = "patient"` so the same grouped-layout machinery
puts P28-only clones on one side, shared clones in the middle, and P32-only
clones on the other side.

```{r gliph_plot, fig.height=7, fig.width=10}
set.seed(seed.number)

p_gliph <- plot_repertoire_network(
  g_gliph,
  color_by    = "patient",
  color_pal   = pub_palettes$patient,
  layout      = "grouped",
  layout_args = list(compartment_col = "patient",
                     csf_label       = "28",
                     pb_label        = "32",
                     node_radius     = 0.025,
                     gap_factor      = 1.25,
                     aspect          = 1.4),
  edge_alpha  = 0.45,
  edge_width  = 0.30,
  size_range  = c(1.2, 6.5),
  label_top_n = 10,
  label_col   = "net_node_label",
  band_labels = FALSE,
  title       = "GLIPH2 motif network — highly expanded CSF clones",
  subtitle    = "Disease-only clones (left)   Shared motif groups (centre)   Healthy-only clones (right)",
  seed        = seed.number
)
p_gliph
```

```{r gliph_pdf, eval=FALSE}
save_pub_figure(p_gliph,
                file_basename = file.path(fig_dir, "Fig4_GLIPH_motif_network"),
                width_class   = "double",
                height_mm     = 120,
                also_png      = TRUE)
```

### -----------------------------

# 6. V-J gene pairing chord diagrams (Figure 5)

A chord diagram shows which V genes pair with which J genes at what frequency,
making compartment- or status-specific biases visually obvious. We render one
per (cell type, compartment) cell.

### a. TCR beta V-J pairing — Disease CSF vs PB

```{r vj_tcr_chord, fig.height=4, fig.width=8}
tcr_beta <- tcr_df[tcr_df$locus %in% "TRB" & tcr_df$patient %in% "28", ]
tcr_beta_csf <- tcr_beta[tcr_beta$COMPARTMENT %in% "CSF", ]
tcr_beta_pb  <- tcr_beta[tcr_beta$COMPARTMENT %in% "PB",  ]

par(mfrow = c(1, 2))
plot_vj_chord(tcr_beta_csf, top_v = 12, top_j = 8,
              v_color = pub_palettes$compartment[["CSF"]],
              j_color = pub_palettes$compartment[["PB"]],
              title = "Disease CSF TCR-β V-J pairing")
plot_vj_chord(tcr_beta_pb,  top_v = 12, top_j = 8,
              v_color = pub_palettes$compartment[["CSF"]],
              j_color = pub_palettes$compartment[["PB"]],
              title = "Disease PB  TCR-β V-J pairing")
par(mfrow = c(1, 1))
```

### b. Save vector PDFs

```{r vj_chord_pdf, eval=FALSE}
plot_vj_chord(tcr_beta_csf, top_v = 12, top_j = 8,
              pdf_out = file.path(fig_dir, "Fig5a_VJ_TRB_CSF.pdf"),
              title = "Disease CSF TCR-β V-J pairing")
plot_vj_chord(tcr_beta_pb,  top_v = 12, top_j = 8,
              pdf_out = file.path(fig_dir, "Fig5b_VJ_TRB_PB.pdf"),
              title = "Disease PB  TCR-β V-J pairing")
```

### c. BCR heavy-chain V-J pairing

```{r vj_bcr_chord, fig.height=4, fig.width=8}
bcr_heavy_dis <- bcr_df[bcr_df$locus %in% "IGH" & bcr_df$patient %in% "28", ]

par(mfrow = c(1, 2))
plot_vj_chord(bcr_heavy_dis[bcr_heavy_dis$COMPARTMENT %in% "CSF", ],
              top_v = 15, top_j = 6, title = "Disease CSF IGH V-J pairing")
plot_vj_chord(bcr_heavy_dis[bcr_heavy_dis$COMPARTMENT %in% "PB", ],
              top_v = 15, top_j = 6, title = "Disease PB  IGH V-J pairing")
par(mfrow = c(1, 1))
```

### -----------------------------

# 7. Clone-size rank distribution (Figure 6)

A Zipf-style log-log plot of clone rank vs clone size. A straight line of
slope ≈ -1 signals a power-law repertoire (typical of antigen-experienced T
cells). The disease CSF should sit visibly higher and flatter than healthy.

### a. Per-compartment, per-patient

```{r rank_curves, fig.height=4, fig.width=10}
# Build a per-(patient × compartment) long table of clone sizes. Each row is
# one clone within a (patient, compartment) group; the cell_count column is
# the number of distinct cells supporting that clone in that group.
clone_counts_long <- tcr_df %>%
  dplyr::distinct(unique_cell_id, clone_id, patient, COMPARTMENT) %>%
  dplyr::group_by(patient, COMPARTMENT, clone_id) %>%
  dplyr::summarize(cell_count = dplyr::n_distinct(unique_cell_id),
                   .groups = "drop") %>%
  dplyr::mutate(group = paste(patient, COMPARTMENT, sep = "_"))

# Color: blue family = disease, orange family = healthy
group_pal <- c(`28_CSF` = "#08306B", `28_PB` = "#6BAED6",
               `32_CSF` = "#7F2704", `32_PB` = "#FDAE6B")

p_rank <- plot_clone_rank(
  clone_count_df = clone_counts_long,
  group_col      = "group",
  count_col      = "cell_count",
  color_pal      = group_pal,
  min_count      = 1,
  title          = "TCR clone-size rank distribution",
  subtitle       = "Zipf-style log-log; disease CSF clones flatter & higher than PB"
)
p_rank
```

```{r rank_pdf, eval=FALSE}
save_pub_figure(p_rank,
                file_basename = file.path(fig_dir, "Fig6_TCR_rank_distribution"),
                width_class   = "1.5", height_mm = 80, also_png = TRUE)
```

### -----------------------------

# 8. Hill diversity profiles (Figure 7)

Hill numbers (q-D) interpolate smoothly from clone richness (q=0) through
Shannon (q=1) to Simpson (q=2) and beyond. A diversity profile is the most
informative single-panel summary of repertoire structure, and is preferred
over a single index in current immunology papers.

```{r hill, fig.height=4, fig.width=10}
# Build a per-(patient × compartment) group column in tcr_df / bcr_df
tcr_df$group <- paste(tcr_df$patient, tcr_df$COMPARTMENT, sep = "_")
bcr_df$group <- paste(bcr_df$patient, bcr_df$COMPARTMENT, sep = "_")

p_hill_tcr <- plot_hill_diversity(
  tcr_df[tcr_df$locus %in% "TRB", ],
  group_col = "group", color_pal = group_pal,
  nboot     = 100,
  title     = "TCR-β Hill diversity",
  subtitle  = "Bootstrap 95% CI; q=0 richness → q=4 dominance"
)

p_hill_bcr <- plot_hill_diversity(
  bcr_df[bcr_df$locus %in% "IGH", ],
  group_col = "group", color_pal = group_pal,
  nboot     = 100,
  title     = "BCR-heavy Hill diversity",
  subtitle  = "Bootstrap 95% CI; q=0 richness → q=4 dominance"
)

fig_hill <- (p_hill_tcr | p_hill_bcr) +
  plot_annotation(tag_levels = "a") +
  plot_layout(guides = "collect")
fig_hill
```

```{r hill_pdf, eval=FALSE}
save_pub_figure(fig_hill,
                file_basename = file.path(fig_dir, "Fig7_Hill_diversity"),
                width_class   = "double", height_mm = 85, also_png = TRUE)
```

### -----------------------------

# 9. Master multi-panel figure (Figure 8 — main figure candidate)

A single master figure combining the disease network, the GLIPH motif network,
the rank distribution and the Hill profile. This is the kind of composite
figure that would land as Figure 2 or 3 of an immunology paper.

```{r master, fig.height=11, fig.width=14}
fig_master <- (
  (p_bcr_dis | p_tcr_dis) /
  (p_gliph   | p_rank)    /
  (p_hill_tcr | p_hill_bcr)
) +
  plot_annotation(tag_levels = "a",
                  title    = "Multi-modal repertoire analysis — disease vs healthy CSF",
                  subtitle = "BCR networks, GLIPH motifs, Zipf rank, Hill diversity") &
  theme(plot.tag = element_text(face = "bold", size = 13))
fig_master
```

```{r master_pdf, eval=FALSE}
save_pub_figure(fig_master,
                file_basename = file.path(fig_dir, "Fig8_master_repertoire"),
                width_class   = "double", height_mm = 240, also_png = TRUE)
```

### -----------------------------

# 10. Session info

```{r session_info}
sessionInfo()
```
