---
title: "VDJ Repertoire Networks"
author: "BMS 270 Computational Immunology Workshop"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  markdown:
    wrap: 78
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  dev        = "cairo_pdf"
)
```

## Overview

This script takes the BCR / TCR repertoire data already produced by
**1_filter_rnaseq_and_vdj_data.Rmd** through **4_find_candidates.Rmd** and
re-renders the network figures at publication quality. The goal is figures
that could go straight into a high-impact paper without further re-styling.

What this script adds compared to `4_find_candidates.Rmd`:

1.  **ggraph + tidygraph backend.** Networks are drawn through ggplot2's
    grammar, so layout, color, size, theme and legend all behave like any
    other ggplot.
2.  **Consistent palettes.** Compartment and patient colors match the base-R
    networks in `4_find_candidates.Rmd` (CSF / P28 = dark blue, PB / P32 =
    tomato); continuous variables use a colorblind-safe viridis ramp.
3.  **Multi-panel composition with patchwork.** Each Figure X is a single
    object you can write to a single PDF.
4.  **Companion visualizations** that contextualize the networks:
    - V-J gene-pairing chord diagrams (`circlize`)
    - Clone-size rank distributions (Zipf-style log-log plots)
    - Hill diversity profiles (q = 0 → 4) with bootstrap 95% CIs
5.  **PDF (vector) output at journal column widths** (85 / 115 / 180 mm) using
    the `cairo_pdf` device, ready for Adobe Illustrator / Inkscape touch-up.

### Prior to running this code

- Set the working directory to the **CompImmunologyWorkshopUCSF** folder:
  `setwd("[PATH]/CompImmunologyWorkshopUCSF")`
- This script depends on the same processed objects as `4_find_candidates.Rmd`
  (i.e. `objects/processed_objects.RData`). It is otherwise self-contained.

### Additional packages required

```{r install_check, eval=FALSE}
# Run once to make sure the publication-figure stack is installed
pkgs <- c("tidygraph", "ggraph", "graphlayouts", "patchwork", "circlize",
          "ggrepel", "ggnewscale", "scales", "ggsci", "tibble", "tidyr")
to_install <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, repos = "https://cloud.r-project.org")
```

```{r libs, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  # Core data
  library(dplyr); library(tidyr); library(tibble); library(stringr)
  library(data.table)
  # Network / graph stack
  library(igraph); library(tidygraph); library(ggraph)
  # Plotting
  library(ggplot2); library(scales); library(patchwork)
  library(RColorBrewer); library(ggrepel); library(ggnewscale)
  # Existing workshop helpers
  library(impactSingleCellToolkit)
})

source("resources/functions_workshop.R")
source("resources/functions_pub_networks.R")

# Reproducibility
seed.number <- 1234
set.seed(seed.number)

# Where to write figures
fig_dir <- "figures"
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

# 1. Load processed objects

```{r load_data}
load("objects/processed_objects.RData")

# Add patient / status columns (mirrors 4_find_candidates.Rmd)
bcr_df$patient <- tstrsplit(bcr_df$sample, "_")[[1]]
tcr_df$patient <- tstrsplit(tcr_df$sample, "_")[[1]]

# Friendly labels
patient_label <- c(`28` = "Disease (P28)", `32` = "Healthy (P32)")
bcr_df$patient_label <- patient_label[bcr_df$patient]
tcr_df$patient_label <- patient_label[tcr_df$patient]
```

### -----------------------------

# 2. BCR repertoire networks (Figure 1)

We build one tidygraph per patient and lay them out side-by-side. Each node is
a B cell; each cluster of connected nodes is a clonotype. Node size is
log-scaled to the clone size; fill encodes compartment (CSF vs PB).

The layout is **category-constrained**: each clone is classified as CSF-pure,
compartment-shared, or PB-pure based on the composition of its member cells,
and clones are sorted into side-by-side bands accordingly. Bands that contain
no clones are dropped — if a repertoire has no compartment-shared clones, the
figure is drawn as two bands (CSF | PB) with no blank middle strip. Within
each band the clones are laid out with `igraph::layout_components()`: every
clone gets an organic Fruchterman-Reingold layout and the clones are packed
so they never overlap, which gives the panel a natural "real network" look
rather than a rigid grid. Band widths are matched to the aspect ratio of
each band's packed layout, so the rescale into the panel is distortion-free.
The result is a network where the spatial story is readable at a glance:
"look at the left band → those are CSF-restricted clones".

### a. Build nodes / edges / graphs

```{r bcr_graphs}
# Use the workshop helpers (same as 4_find_candidates.Rmd) so the underlying
# node/edge structure is identical -- we just re-render it nicely.
bcr_node_dis <- createRepertoireNodeDataframe(
  bcr_df, assay = "bcr", select.patients = "28", expansion.thresh = 2
)
# id format from the helper is "{barcode}-{patient}_{compartment}", with
# compartment encoded as PBMC. Normalize to PB to match the rest of the data.
bcr_node_dis$compartment <- normalize_compartment(tstrsplit(bcr_node_dis$id, "_")[[2]])
bcr_node_dis$patient     <- "28"
bcr_edges_dis <- createRepertoireEdgeDataframe(bcr_node_dis)
g_bcr_dis     <- build_repertoire_tidygraph(bcr_node_dis, bcr_edges_dis)

bcr_node_hc <- createRepertoireNodeDataframe(
  bcr_df, assay = "bcr", select.patients = "32", expansion.thresh = 2
)
bcr_node_hc$compartment <- normalize_compartment(tstrsplit(bcr_node_hc$id, "_")[[2]])
bcr_node_hc$patient     <- "32"
bcr_edges_hc <- createRepertoireEdgeDataframe(bcr_node_hc)
g_bcr_hc     <- build_repertoire_tidygraph(bcr_node_hc, bcr_edges_hc)
```

### b. Render each panel

The CSF / PB band captions are drawn inside each panel, so the panel
subtitle is kept short to avoid the title block crowding on PDF export.

```{r bcr_plots, fig.height=6, fig.width=13}
# Shared styling for both BCR panels. `band_fill` (< 1) leaves a margin so
# clones never touch the band edge; `aspect` is the overall canvas shape.
bcr_layout_args <- list(aspect = 1.7, band_fill = 0.90, band_gap = 0.10)

# A shared clone-size limit so patchwork collects ONE size legend for the
# pair instead of a slightly different legend per panel.
bcr_size_lim <- range(c(as_tibble(g_bcr_dis, "nodes")$clone_size,
                        as_tibble(g_bcr_hc,  "nodes")$clone_size),
                      na.rm = TRUE)

p_bcr_dis <- plot_repertoire_network(
  g_bcr_dis,
  color_by    = "compartment",
  layout      = "grouped",
  layout_args = bcr_layout_args,
  edge_alpha  = 0.30,
  edge_width  = 0.30,
  size_range  = c(1.2, 4.5),
  size_limits = bcr_size_lim,
  label_top_n = 5,
  label_col   = "clone_id",
  band_labels = TRUE,
  title       = "B cell repertoire — Disease (P28)",
  subtitle    = "Clones with ≥ 2 cells; fill = compartment",
  seed        = seed.number
)

p_bcr_hc <- plot_repertoire_network(
  g_bcr_hc,
  color_by    = "compartment",
  layout      = "grouped",
  layout_args = bcr_layout_args,
  edge_alpha  = 0.30,
  edge_width  = 0.30,
  size_range  = c(1.2, 4.5),
  size_limits = bcr_size_lim,
  label_top_n = 5,
  label_col   = "clone_id",
  band_labels = TRUE,
  title       = "B cell repertoire — Healthy (P32)",
  subtitle    = "Clones with ≥ 2 cells; fill = compartment",
  seed        = seed.number
)

fig_bcr <- compose_panels(list(p_bcr_dis, p_bcr_hc), ncol = 1,
                          tag_levels = "a")
fig_bcr
```

### c. Save vector PDF + 300-DPI PNG

```{r bcr_pdf, eval=FALSE}
save_pub_figure(fig_bcr,
                file_basename = file.path(fig_dir, "BCR_networks"),
                width_class   = "double",
                height_mm     = 130,
                also_png      = TRUE)
```

### -----------------------------

# 3. TCR (expanded) repertoire networks (Figure 2)

T cell repertoires are much larger than BCR repertoires, so we restrict to
expanded clones (cell count \> 1) and split by CD4 vs CD8 in the next panel.

### a. Build graphs

```{r tcr_graphs}
tcr_df_exp <- tcr_df[! tcr_df$EXPANSION_CATEGORY %in% "Non-expanded", ]

# Disease
tcr_node_dis <- createRepertoireNodeDataframe(
  tcr_df_exp, assay = "tcr", select.patients = "28", expansion.thresh = 2
)
tcr_node_dis$compartment <- normalize_compartment(tstrsplit(tcr_node_dis$id, "_")[[2]])
tcr_node_dis$patient     <- "28"
tcr_edges_dis <- createRepertoireEdgeDataframe(tcr_node_dis)
g_tcr_dis     <- build_repertoire_tidygraph(tcr_node_dis, tcr_edges_dis)

# Healthy
tcr_node_hc <- createRepertoireNodeDataframe(
  tcr_df_exp, assay = "tcr", select.patients = "32", expansion.thresh = 2
)
tcr_node_hc$compartment <- normalize_compartment(tstrsplit(tcr_node_hc$id, "_")[[2]])
tcr_node_hc$patient     <- "32"
tcr_edges_hc <- createRepertoireEdgeDataframe(tcr_node_hc)
g_tcr_hc     <- build_repertoire_tidygraph(tcr_node_hc, tcr_edges_hc)
```

### b. Render

```{r tcr_plots, fig.height=6, fig.width=13}
# T cell repertoires have many more clones, so the canvas is a touch wider
# and the nodes a touch smaller.
tcr_layout_args <- list(aspect = 1.9, band_fill = 0.92, band_gap = 0.10)

tcr_size_lim <- range(c(as_tibble(g_tcr_dis, "nodes")$clone_size,
                        as_tibble(g_tcr_hc,  "nodes")$clone_size),
                      na.rm = TRUE)

p_tcr_dis <- plot_repertoire_network(
  g_tcr_dis,
  color_by    = "compartment",
  layout      = "grouped",
  layout_args = tcr_layout_args,
  edge_alpha  = 0.15,
  edge_width  = 0.20,
  size_range  = c(0.8, 3.6),
  size_limits = tcr_size_lim,
  label_top_n = 4,
  band_labels = TRUE,
  title       = "T cell repertoire — Disease (P28)",
  subtitle    = "Expanded clones only (cell count > 1)",
  seed        = seed.number
)

p_tcr_hc <- plot_repertoire_network(
  g_tcr_hc,
  color_by    = "compartment",
  layout      = "grouped",
  layout_args = tcr_layout_args,
  edge_alpha  = 0.15,
  edge_width  = 0.20,
  size_range  = c(0.8, 3.6),
  size_limits = tcr_size_lim,
  label_top_n = 4,
  band_labels = TRUE,
  title       = "T cell repertoire — Healthy (P32)",
  subtitle    = "Expanded clones only (cell count > 1)",
  seed        = seed.number
)

fig_tcr <- compose_panels(list(p_tcr_dis, p_tcr_hc), ncol = 1, tag_levels = "a")
fig_tcr
```

```{r tcr_pdf, eval=FALSE}
save_pub_figure(fig_tcr,
                file_basename = file.path(fig_dir, "TCR_networks"),
                width_class   = "double",
                height_mm     = 130,
                also_png      = TRUE)
```


### -----------------------------

# 4. GLIPH motif network (Figure 4)

GLIPH2 groups TCRs by shared CDR3 motifs that putatively bind the same
antigen. Here we re-render the GLIPH network with proper labels on the motif
hubs, colored by patient, sized by clone size.

### a. Re-run the same filtering as `4_find_candidates.Rmd`

```{r gliph_filter}
gliph_results_fh <- "4_FindCandidates/all_CSF_GLIPH2/AllCSF_cluster.csv"
gliph_resuls_csf <- read.csv(gliph_results_fh, header = TRUE, stringsAsFactors = FALSE)
gliph_resuls_csf$CRD3ab_AA <- paste(gliph_resuls_csf$TcRa, gliph_resuls_csf$TcRb, sep = ":")

tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_df,
                                                   clone_counts = clone_counts_tcr,
                                                   target.col = "pct_CSF",
                                                   new.col.name = "clonotype_CSF_freq")
tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_results_gliph,
                                                   clone_counts = clone_counts_tcr,
                                                   target.col = "cell_count",
                                                   new.col.name = "cell_count")
tcr_results_gliph <- addCloneCountColToVDJresults(results = tcr_results_gliph,
                                                   clone_counts = clone_counts_tcr,
                                                   target.col = "CSF_count",
                                                   new.col.name = "CSF_count")

# Build CDR3ab_AA per cell
tcr_cdr3_col <- tcr_results_gliph$unique_cell_id
for (cell in unique(tcr_results_gliph$unique_cell_id)) {
  cell_df <- tcr_results_gliph[tcr_results_gliph$unique_cell_id %in% cell, ]
  alpha   <- cell_df[cell_df$locus %in% "TRA", ]
  beta    <- cell_df[cell_df$locus %in% "TRB", ]
  ab      <- paste(c(alpha$junction_aa, beta$junction_aa), collapse = ":")
  tcr_cdr3_col[tcr_cdr3_col %in% cell] <- ab
}
tcr_results_gliph$CDR3ab_AA <- tcr_cdr3_col

# GLIPH membership flag
gliph_status_col <- tcr_results_gliph$CDR3ab_AA
gliph_status_col[gliph_status_col %in% unique(gliph_resuls_csf$CRD3ab_AA)] <- "1"
gliph_status_col[! gliph_status_col %in% "1"] <- "0"
tcr_results_gliph$gliph_status <- gliph_status_col

# Index group string per cell
target_cells <- tcr_results_gliph[tcr_results_gliph$CDR3ab_AA %in% gliph_resuls_csf$CRD3ab_AA, ]
target_cells <- unique(target_cells$unique_cell_id)

index_col        <- tcr_results_gliph$CDR3ab_AA
names(index_col) <- tcr_results_gliph$unique_cell_id
index_col[! names(index_col) %in% target_cells] <- ""
target_cdr3ab <- unique(as.character(index_col[! index_col %in% ""]))
for (ab in target_cdr3ab) {
  gliph_sub <- gliph_resuls_csf[gliph_resuls_csf$CRD3ab_AA %in% ab, ]
  if (nrow(gliph_sub) > 0) {
    index_vec <- unique(gliph_sub$index)
    if (length(index_vec) > 1) {
      index_col[index_col %in% ab] <- paste(sort(index_vec), collapse = ":")
    } else {
      index_col[index_col %in% ab] <- index_vec
    }
  }
}
tcr_results_gliph$gliph_group <- index_col

# Subset to expanded, GLIPH-positive CSF clones (mirrors 4_find_candidates.Rmd)
THRESH <- 0.75
clonotype_counts_csf     <- clone_counts_tcr[clone_counts_tcr$CSF_count > 1, ]
clonotype_counts_csf_exp <- clonotype_counts_csf[clonotype_counts_csf$cell_count > 1, ]
high_csf_clones <- clonotype_counts_csf_exp[clonotype_counts_csf_exp$CSF_PB_ratio > THRESH, ]
target_clones   <- row.names(high_csf_clones)

data_tcr_na     <- tcr_results_gliph[is.na(tcr_results_gliph$junction_aa), ]
incomplete_tcrs <- unique(data_tcr_na$unique_cell_id)
data_tcr_clean  <- tcr_results_gliph[! tcr_results_gliph$unique_cell_id %in% incomplete_tcrs, ]

data_tcr_csf     <- data_tcr_clean[data_tcr_clean$COMPARTMENT %in% "CSF", ]
data_tcr_csf_exp <- data_tcr_csf[data_tcr_csf$clone_id %in% row.names(clonotype_counts_csf_exp), ]
data_tcr_csf_exp <- data_tcr_csf_exp[data_tcr_csf_exp$gliph_status %in% "1", ]
data_tcr_target  <- data_tcr_csf_exp[data_tcr_csf_exp$clone_id %in% target_clones, ]

target_groups <- unique(data_tcr_target$gliph_group)
target_groups_all <- unique(unlist(strsplit(target_groups, ":")))
target_groups_all <- c(target_groups_all, target_groups)
target_groups_all <- unique(target_groups_all)

interac_tcr <- unique(data_tcr_csf[data_tcr_csf$gliph_group %in% target_groups_all,
                                    "unique_cell_id"])

all_target_cells     <- c(unique(data_tcr_target$unique_cell_id), interac_tcr)
data_tcr_target_all  <- data_tcr_clean[data_tcr_clean$unique_cell_id %in% all_target_cells, ]
data_tcr_target_all  <- data_tcr_target_all[data_tcr_target_all$gliph_status %in% "1", ]
```

### b. Build GLIPH-grouped node / edge / graph

```{r gliph_graph}
gliph_node <- createRepertoireNodeDataframe(
  data_tcr_target_all, assay = "tcr",
  clone.col       = "gliph_group",
  prefix.clone.id = "gliph",
  expansion.thresh = 2
)
row.names(gliph_node) <- gliph_node$id

# Add patient and clone-id metadata to nodes (one row per cell)
extra_df <- data_tcr_target_all[data_tcr_target_all$locus %in% "TRB", ]
extra_df <- extra_df[!duplicated(extra_df$unique_cell_id), ]
row.names(extra_df) <- extra_df$unique_cell_id
extra_df <- extra_df[, c("clone_id", "patient", "COMPARTMENT")]
names(extra_df)[1] <- "clone_id_vdj"

gliph_node <- merge(gliph_node, extra_df, by = 0, all.x = TRUE)
gliph_node[, 1] <- NULL
row.names(gliph_node) <- gliph_node$id

# Compact motif label: one label per GLIPH group, on the largest node
net_label_col <- gliph_node$clone_id
for (grp in unique(net_label_col)) {
  idx <- which(net_label_col %in% grp)
  # keep the label on the first row (one motif label per cluster is enough)
  if (length(idx) > 1) net_label_col[idx[-1]] <- ""
}
gliph_node$net_node_label <- net_label_col

gliph_edges <- createRepertoireEdgeDataframe(gliph_node)
g_gliph     <- build_repertoire_tidygraph(gliph_node, gliph_edges)
```

### c. Render — labels on motif hubs, grouped layout by patient

For the GLIPH network the meaningful spatial axis is *patient identity*
rather than compartment, since GLIPH groups are by definition CSF-only here.
We use `compartment_col = "patient"` so the same grouped-layout machinery
puts P28-only clones in one band, shared GLIPH groups in the middle, and
P32-only clones in the other band. `band_titles` re-labels the bands in
patient terms, and the patient palette (P28 = dark blue, P32 = tomato)
matches `4_find_candidates.Rmd`. If there are no patient-shared GLIPH groups
the middle band is dropped automatically.

```{r gliph_plot, fig.height=7, fig.width=10}
set.seed(seed.number)

p_gliph <- plot_repertoire_network(
  g_gliph,
  color_by    = "patient",
  color_pal   = pub_palettes$patient,
  layout      = "grouped",
  layout_args = list(compartment_col = "patient",
                     csf_label       = "28",
                     pb_label        = "32",
                     aspect          = 1.6,
                     band_fill       = 0.92,
                     band_gap        = 0.10),
  edge_alpha  = 0.25,
  edge_width  = 0.22,
  size_range  = c(1.0, 4.8),
  label_top_n = 10,
  label_col   = "net_node_label",
  band_labels = TRUE,
  band_titles = c(CSF   = "Disease-only (P28)",
                  Mixed = "Shared motifs",
                  PB    = "Healthy-only (P32)"),
  title       = "GLIPH2 motif network — highly expanded CSF clones",
  subtitle    = "GLIPH groups; fill = patient",
  seed        = seed.number
)
p_gliph
```

```{r gliph_pdf, eval=FALSE}
save_pub_figure(p_gliph,
                file_basename = file.path(fig_dir, "GLIPH_motif_TCR_network"),
                width_class   = "double",
                height_mm     = 120,
                also_png      = TRUE)
```


### -----------------------------

# Session info

```{r session_info}
sessionInfo()
```
